---
title: "Batch Processing with root_depth_metrics()"
author: "Johannes Cunow"
output:
  rmarkdown::html_vignette: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Batch Processing with root_depth_metrics()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, dpi = 300, fig.width = 9)
```

# Batch Processing with `root_depth_metrics()`

## Who this vignette is for

If you have a folder of segmented minirhizotron images and want depth-resolved
root traits for all of them in one call — without writing loops yourself —
`root_depth_metrics()` is the right starting point.

It handles file loading, depth-map construction, per-bin zonal statistics,
derived metrics, fault tolerance, and optional file output. You toggle which
metric groups to compute; everything else is taken care of.

For a line-by-line explanation of the underlying steps, see the
[Minirhizotron Scans](MinirhizotronScans_vignettes.html) vignette.
If multiple images come from the same tube, consider stitching them beforehand [Image Stitching](Stitching_vignette.html)

---

## Minimum requirements

| What you need | Where it comes from |
|---|---|
| Segmented images (binary, root = 1) | [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or [RootPainter](https://github.com/Abe404/root_painter) |
| Skeletonised images (one-pixel centerlines) — *optional* | RootDetector (channel 2), or omit `path_skl` and `root_depth_metrics()` will compute skeletons internally via `skeletonize_image()` — needed for length, diameter, angle, and branching-order metrics |
| RGB images aligned to segmented images | Original scan — needed for color metrics only |
| Tube insertion angle per image | Field metadata |
| Soil-surface pixel row per image — *optional*, defaults to `0` | Field metadata or `estimate_soil_surface()`; only needed if you want depths reported relative to the true soil surface rather than the top of the image |

Images in the three directories must be in the **same alphabetical order** as
each other; each position corresponds to the same physical tube/window.

---

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("jcunow/Rootopia")

library(Rootopia)
library(tidyverse)   # for downstream plotting
```

---

## Quick start — default metrics only

This is the fastest way to get going. Only `path_seg` is required; everything
else has sensible defaults. `path_skl` is optional — if omitted, skeletons
needed for length, diameter, and angle metrics are computed internally via
`skeletonize_image()`.

```{r quickstart}
result <- root_depth_metrics(
  path_seg = "scans/segmented/2022_02/",
  path_skl = "scans/skeleton/2022_02/",   # optional — computed internally if omitted
  session  = "2022_02"
)

head(result)
```

The result is a data frame with one row per tube × depth-bin combination.
Always-present columns: `Tube`, `Session`, `Plot`, `depth`.
Default metric columns: `rootpx`, `voidpx`, `rootlength`, `avg.diameter`,
`max.diameter`, `var.diameter`, `root.surface.area`, `root.volume`,
`rootpx.density`, `rootlength.density`,
`mrd`, `total.length.density`, `rootlength.fraction`,
`mean.var.diameter`, `rootsurface_rootvolume_ratio`.

---

## Using per-image metadata

When tubes differ in insertion angle or soil-surface position, supply a vector
of values (one per image). Both accept a single recycled value when all images
share the same setting.

```{r metadata}
# Load your field metadata table
tube_meta <- read.csv("metadata/tube_metadata_2022_02.csv")
# Expected columns: filename, angle, soil_row, tube_id

result <- root_depth_metrics(
  path_seg         = "scans/segmented/2022_02/",
  path_skl         = "scans/skeleton/2022_02/",
  insertion_angles = tube_meta$angle,       # degrees from vertical
  soil_starts      = tube_meta$soil_row,    # pixel row of soil surface
  tube_names       = tube_meta$tube_id,     # e.g. "T042"
  session          = "2022_02",
  dpi              = 300
)
```

If your tube names follow a different convention from the default (characters
3–5 from the right of the filename), always supply `tube_names` explicitly.

---

## Toggling metric groups

Each metric group is a logical argument. Disabled groups produce `NA` columns
rather than absent columns, so your downstream code can always reference the
same column names regardless of what was computed.

```{r metrics}
result <- root_depth_metrics(
  path_seg  = "scans/segmented/2022_02/",
  path_skl  = "scans/skeleton/2022_02/",
  path_rgb  = "scans/blended/2022_02/",
  session   = "2022_02",

  # --- core (on by default, shown explicitly for clarity) ---
  calc_root_pixels    = TRUE,
  calc_root_length    = TRUE,
  calc_diameter_stats = TRUE,

  # --- derived (on by default) ---
  calc_density_metrics      = TRUE,
  calc_distribution_indices = TRUE,
  calc_advanced_metrics     = TRUE,

  # --- extended (off by default — enable as needed) ---
  calc_diameter_quantiles = TRUE,   # 90th/95th/99th percentile + modal peaks
  calc_color_metrics      = TRUE,   # RGB color per root vs background pixels
  calc_root_angles        = TRUE,   # deep_drive + steepness angle distribution
  calc_root_order_metrics = TRUE,   # branching order per segment
  calc_landscape_metrics  = FALSE   # slow — enable only when needed
)
```

### What each group adds

| Argument | New columns | Notes |
|---|---|---|
| `calc_root_pixels` | `rootpx`, `voidpx` | Needed for density metrics |
| `calc_root_length` | `rootlength` | Needed for density + angle metrics |
| `calc_diameter_stats` | `avg.diameter`, `max.diameter`, `var.diameter`, `root.surface.area`, `root.volume`, `rootsurface_rootvolume_ratio` | Surface = lateral (soil-contact) area of the cylindrical root, not the image-visible area. Ratio is the length-weighted mean of local 2/r (fine-root dominated), not `root.surface.area / root.volume` |
| `calc_diameter_quantiles` | `rootdiameter.90/95/99`, `avg.diameter.top*pct`, `rootlength.above.*`, `n.diameter.peaks`, `diameter.peak.*` | See `diameter_quantiles`, `diameter_thresholds` |
| `calc_landscape_metrics` | `enn_mn`, `joinent`, `relmutinf`, `np`, `contag`, `np_density` | **Slow** — one call per depth bin |
| `calc_color_metrics` | `rcc_root`, `gcc_root`, ..., `rcc_bg`, `gcc_bg`, ... | Requires `path_rgb` |
| `calc_root_angles` | `deep_drive`, `mean.steepness.angle`, `sd.steepness.angle` | Requires `calc_root_length` |
| `calc_root_order_metrics` | Per bin: `mean.branch_order`, `max.branch_order`, `mean.root_order`, `lateral_root_fraction`. Per tube: `main_root.*` / `lateral_roots.*` (length, diameter, branching frequency, ...), `n_root_orders` | **Slow** — builds one segment graph per image. Requires `calc_root_length` |
| `calc_density_metrics` | `rootpx.density`, `rootlength.density` | Auto-enables pixels + length |
| `calc_distribution_indices` | `mrd`, `total.length.density` | Tube-level summary |
| `calc_advanced_metrics` | `rootlength.fraction`, `mean.var.diameter` | Auto-enables distribution indices + diameter stats |

---

## Diameter thresholds

When `calc_diameter_quantiles = TRUE`, you can specify custom diameter
thresholds that separate fine roots from coarse roots:

```{r thresholds}
result <- root_depth_metrics(
  path_seg                = "scans/segmented/2022_02/",
  path_skl                = "scans/skeleton/2022_02/",
  calc_diameter_quantiles = TRUE,
  diameter_thresholds     = c(0.2, 0.5, 1.0),   # fine / medium / coarse
  diameter_threshold_unit = "mm"                 # "mm", "cm", or "px"
)
# Resulting columns include: rootlength.above.0.2mm, rootlength.above.0.5mm, ...
```

---

## Root branching order

`calc_root_order_metrics = TRUE` builds a segment graph from the skeleton of
each image (via `branch_order_map()`) and assigns every root segment a
**branching order**: segments on the thickest, most central path of each
connected component are order 1, the branches coming off them are order 2,
their branches order 3, and so on. See the
[Flatbed Scans](FlatBedScans_vignettes.html#7b-branch-order)
vignette for a worked example of the underlying pipeline.

```{r order}
result <- root_depth_metrics(
  path_seg                = "scans/segmented/2022_02/",
  path_skl                = "scans/skeleton/2022_02/",
  session                 = "2022_02",
  calc_root_order_metrics = TRUE
)
```

This adds two kinds of columns:

**Per depth bin** — how branching order varies with depth:

| Column | Meaning |
|---|---|
| `mean.branch_order` | Mean branch order of skeleton pixels in the bin |
| `max.branch_order` | Highest branch order reached in the bin |
| `mean.root_order` | Mean root order (max tip-order along each root) in the bin |
| `lateral_root_fraction` | Fraction of skeleton pixels belonging to branch order > 1 (laterals) |

**Per tube** (repeated on every row of that tube) — the thickest order-1
path vs. all higher-order branches, split by
`order_metrics(..., focal = "thickest")`:

| Column prefix | Meaning |
|---|---|
| `main_root.*` | Totals/means for the thickest (order-1) root(s): `total_length`, `mean_diameter`, `median_diameter`, `n_segments`, `n_tips`, `n_branch_points`, `length_fraction`, `mean_segment_length`, `branching_frequency` |
| `lateral_roots.*` | The same set of statistics for everything else (order >= 2) |
| `n_root_orders` | Highest branch order found in the image |

```{r order-plot}
library(ggplot2)

ggplot(result, aes(x = depth, y = lateral_root_fraction, color = Tube)) +
  geom_line() +
  geom_point() +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(
    title = "Lateral Root Fraction by Depth",
    x     = "Soil Depth (cm)",
    y     = "Fraction of skeleton pixels with branch order > 1"
  )
```

> **Slow.** This builds one segment graph per image, which is considerably
> more expensive than the other metric groups. Disable it (the default) for
> quick exploratory runs and enable it only when you need architectural
> detail.

---

## Flat rhizotron windows

For glass-fronted rhizotron boxes (flat geometry, no tube curvature), set
`flat_geometry = TRUE`. This disables the sinusoidal depth correction that
is otherwise applied to cylindrical minirhizotron tubes.

```{r flat}
result <- root_depth_metrics(
  path_seg      = "scans/segmented/rhizotron_A/",
  path_skl      = "scans/skeleton/rhizotron_A/",
  flat_geometry = TRUE,
  insertion_angles = 0   # vertical windows
)
```

---

## Subsetting files

If your skeleton and RGB directories contain more files than the segmented
directory (e.g. multiple sessions in one folder), use `*_file_index` to
select the matching subset:

```{r subset}
# Use only files 37–72 from the RGB directory
result <- root_depth_metrics(
  path_seg       = "scans/segmented/2022_02/",
  path_skl       = "scans/skeleton/2022_02/",
  path_rgb       = "scans/blended/all_sessions/",
  rgb_file_index = 37:72,
  session        = "2022_02"
)
```

---

## Saving output

Pass `output_path` to save the result as an `.RData` file (the object is
named `root.depth.metrics` inside the file). The directory is created if it
does not exist.

```{r save}
result <- root_depth_metrics(
  path_seg    = "scans/segmented/2022_02/",
  path_skl    = "scans/skeleton/2022_02/",
  session     = "2022_02",
  output_path = "output/root_metrics_2022_02.RData"
)

# Reload later:
# load("output/root_metrics_2022_02.RData")
# head(root.depth.metrics)
```

---

## Progress and fault tolerance

By default (`verbose = TRUE`) the function prints a progress line after each
image showing per-image time, cumulative elapsed time, estimated remaining
time, and predicted clock-time of completion:

```
[Rootopia] [3/48] T042 | img: 12s | elapsed: 38s | remaining: ~546s | done ~14:22
```

If a metric block fails for a single image (e.g. a corrupted skeleton file),
it is skipped with a `[Rootopia] SKIPPED` message and the corresponding
columns are filled with `NA`. Processing always continues. Images that cannot
be loaded at all are skipped entirely and listed in a warning at the end.

To suppress all progress output:

```{r quiet}
result <- root_depth_metrics(
  path_seg = "scans/segmented/2022_02/",
  path_skl = "scans/skeleton/2022_02/",
  verbose  = FALSE
)
```

---

## Example: visualizing depth profiles

```{r plot}
library(ggplot2)

# Root length density by depth, one line per tube
ggplot(result, aes(x = depth, y = rootlength.density, color = Tube)) +
  geom_line() +
  geom_point() +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(
    title = "Root Length Density by Depth",
    x     = "Soil Depth (cm)",
    y     = "Root Length Density (cm root / cm² area)"
  )

# Average root diameter by depth
ggplot(result, aes(x = depth, y = avg.diameter)) +
  geom_smooth(method = "loess", se = TRUE) +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(
    title = "Average Root Diameter with Depth",
    x     = "Soil Depth (cm)",
    y     = "Average Diameter (cm)"
  )

# mrd distribution across tubes
ggplot(result |> dplyr::distinct(Tube, mrd), aes(x = mrd)) +
  geom_histogram(bins = 15, fill = "steelblue4", color = "white") +
  theme_minimal() +
  labs(
    title = "Mean Root Depth Distribution (MRD)",
    x     = "MRD",
    y     = "Count"
  )
```

---

## Full example with all options

```{r full}
tube_meta <- read.csv("metadata/tube_metadata_2022_02.csv")

result <- root_depth_metrics(

  # Paths
  path_seg       = "scans/segmented/2022_02/",
  path_skl       = "scans/skeleton/2022_02/",
  path_rgb       = "scans/blended/2022_02/",

  # Per-image metadata
  insertion_angles = tube_meta$angle,
  soil_starts      = tube_meta$soil_row,
  tube_names       = tube_meta$tube_id,
  session          = "2022_02",

  # Geometry
  dpi               = 300,
  tube_diameter_cm  = 7,
  depth_interval_cm = 5,
  flat_geometry     = FALSE,

  # Core metrics
  calc_root_pixels    = TRUE,
  calc_root_length    = TRUE,
  calc_diameter_stats = TRUE,

  # Extended metrics
  calc_diameter_quantiles = TRUE,
  calc_color_metrics      = TRUE,
  calc_root_angles        = TRUE,
  calc_root_order_metrics = TRUE,    # per-segment branching-order architecture
  calc_landscape_metrics  = FALSE,   # enable if you need patch metrics

  # Derived metrics
  calc_density_metrics      = TRUE,
  calc_distribution_indices = TRUE,
  calc_advanced_metrics     = TRUE,

  # Diameter thresholds (for quantile block)
  diameter_thresholds     = c(0.2, 0.5, 1.0),
  diameter_threshold_unit = "mm",

  # Output
  output_path = "output/root_metrics_2022_02.RData",
  verbose     = TRUE
)
```

---

## What to read next

- [Minirhizotron Scans](MinirhizotronScans_vignettes.html) — step-by-step
  explanation of each underlying function
- [Flatbed Scans](FlatBedScans_vignettes.html) — trait extraction from
  flatbed scanner images (no depth dimension)
- [Rotation Bias](Rotation_Bias_vignettes.html) — correcting for tube
  rotation artifacts before analysis
- [Function reference](../reference/index.html) — full documentation for
  every exported function
