---
title: "Batch Processing with root_depth_metrics()"
author: "Johannes Cunow"
output:
  rmarkdown::html_vignette: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Batch Processing with root_depth_metrics()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Batch Processing with `root_depth_metrics()`

## Who this vignette is for

If you have a folder of segmented minirhizotron images and want depth-resolved
root traits for all of them in one call — without writing loops yourself —
`root_depth_metrics()` is the right starting point.

It handles file loading, depth-map construction, per-bin zonal statistics,
derived metrics, fault tolerance, and optional file output. You toggle which
metric groups to compute; everything else is taken care of.

For a line-by-line explanation of the underlying steps, see the
[Minirhizotron Scans](MinirhizotronScans_vignettes.html) vignette.

---

## Minimum requirements

| What you need | Where it comes from |
|---|---|
| Segmented images (binary, root = 1) | [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or [RootPainter](https://github.com/Abe404/root_painter) |
| Skeletonised images (one-pixel centrelines) | RootDetector (channel 2) — needed for length, diameter, angle metrics |
| RGB images aligned to segmented images | Original scan — needed for colour metrics only |
| Tube insertion angle per image | Field metadata |
| Soil-surface pixel row per image | Field metadata or `estimate_soil_surface()` |

Images in the three directories must be in the **same alphabetical order** as
each other; each position corresponds to the same physical tube/window.

---

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("jcunow/RootScanR")

library(RootScanR)
library(tidyverse)   # for downstream plotting
```

---

## Quick start — default metrics only

This is the fastest way to get going. Only `path.seg` and `path.skl` are
required; everything else has sensible defaults.

```{r quickstart}
result <- root_depth_metrics(
  path.seg = "scans/segmented/2022_02/",
  path.skl = "scans/skeleton/2022_02/",
  session  = "2022_02"
)

head(result)
```

The result is a data frame with one row per tube × depth-bin combination.
Always-present columns: `Tube`, `Session`, `Plot`, `depth`.
Default metric columns: `rootpx`, `voidpx`, `rootlength`, `avg.diameter`,
`max.diameter`, `var.diameter`, `rootpx.density`, `rootlength.density`,
`rwdi`, `rpi`, `total.length.density`, `rootlength.fraction`,
`mean.var.diameter`, `rootsurface_rootvolume_ratio`.

---

## Using per-image metadata

When tubes differ in insertion angle or soil-surface position, supply a vector
of values (one per image). Both accept a single recycled value when all images
share the same setting.

```{r metadata}
# Load your field metadata table
tube_meta <- read.csv("metadata/tube_metadata_2022_02.csv")
# Expected columns: filename, angle, soil_row, tube_id

result <- root_depth_metrics(
  path.seg         = "scans/segmented/2022_02/",
  path.skl         = "scans/skeleton/2022_02/",
  insertion_angles = tube_meta$angle,       # degrees from vertical
  soil_starts      = tube_meta$soil_row,    # pixel row of soil surface
  tube_names       = tube_meta$tube_id,     # e.g. "T042"
  session          = "2022_02",
  dpi              = 300
)
```

If your tube names follow a different convention from the default (characters
3–5 from the right of the filename), always supply `tube_names` explicitly.

---

## Toggling metric groups

Each metric group is a logical argument. Disabled groups produce `NA` columns
rather than absent columns, so your downstream code can always reference the
same column names regardless of what was computed.

```{r metrics}
result <- root_depth_metrics(
  path.seg  = "scans/segmented/2022_02/",
  path.skl  = "scans/skeleton/2022_02/",
  path.rgb  = "scans/blended/2022_02/",
  session   = "2022_02",

  # --- core (on by default, shown explicitly for clarity) ---
  calc_root_pixels    = TRUE,
  calc_root_length    = TRUE,
  calc_diameter_stats = TRUE,

  # --- derived (on by default) ---
  calc_density_metrics      = TRUE,
  calc_distribution_indices = TRUE,
  calc_advanced_metrics     = TRUE,

  # --- extended (off by default — enable as needed) ---
  calc_diameter_quantiles = TRUE,   # 90th/95th/99th percentile + modal peaks
  calc_color_metrics      = TRUE,   # RGB colour per root vs background pixels
  calc_root_angles        = TRUE,   # deep_drive + steepness angle distribution
  calc_landscape_metrics  = FALSE   # slow — enable only when needed
)
```

### What each group adds

| Argument | New columns | Notes |
|---|---|---|
| `calc_root_pixels` | `rootpx`, `voidpx` | Needed for density metrics |
| `calc_root_length` | `rootlength` | Needed for density + angle metrics |
| `calc_diameter_stats` | `avg.diameter`, `max.diameter`, `var.diameter` | Needed for surface:volume ratio |
| `calc_diameter_quantiles` | `rootdiameter.90/95/99`, `avg.diameter.top*pct`, `rootlength.above.*`, `n.diameter.peaks`, `diameter.peak.*` | See `diameter_thresholds` |
| `calc_landscape_metrics` | `enn_mn`, `joinent`, `relmutinf`, `np`, `contag`, `np_density` | **Slow** — one call per depth bin |
| `calc_color_metrics` | `rcc_root`, `gcc_root`, ..., `rcc_bg`, `gcc_bg`, ... | Requires `path.rgb` |
| `calc_root_angles` | `deep_drive`, `mean.steepness.angle`, `sd.steepness.angle` | Requires `calc_root_length` |
| `calc_density_metrics` | `rootpx.density`, `rootlength.density` | Auto-enables pixels + length |
| `calc_distribution_indices` | `rwdi`, `rpi`, `total.length.density` | Tube-level summary |
| `calc_advanced_metrics` | `rootlength.fraction`, `mean.var.diameter`, `rootsurface_rootvolume_ratio` | Auto-enables distribution indices + diameter stats |

---

## Diameter thresholds

When `calc_diameter_quantiles = TRUE`, you can specify custom diameter
thresholds that separate fine roots from coarse roots:

```{r thresholds}
result <- root_depth_metrics(
  path.seg                = "scans/segmented/2022_02/",
  path.skl                = "scans/skeleton/2022_02/",
  calc_diameter_quantiles = TRUE,
  diameter_thresholds     = c(0.2, 0.5, 1.0),   # fine / medium / coarse
  diameter_threshold_unit = "mm"                 # "mm", "cm", or "px"
)
# Resulting columns include: rootlength.above.0.2mm, rootlength.above.0.5mm, ...
```

---

## Flat rhizotron windows

For glass-fronted rhizotron boxes (flat geometry, no tube curvature), set
`flat_geometry = TRUE`. This disables the sinusoidal depth correction that
is otherwise applied to cylindrical minirhizotron tubes.

```{r flat}
result <- root_depth_metrics(
  path.seg      = "scans/segmented/rhizotron_A/",
  path.skl      = "scans/skeleton/rhizotron_A/",
  flat_geometry = TRUE,
  insertion_angles = 0   # vertical windows
)
```

---

## Subsetting files

If your skeleton and RGB directories contain more files than the segmented
directory (e.g. multiple sessions in one folder), use `*_file_index` to
select the matching subset:

```{r subset}
# Use only files 37–72 from the RGB directory
result <- root_depth_metrics(
  path.seg       = "scans/segmented/2022_02/",
  path.skl       = "scans/skeleton/2022_02/",
  path.rgb       = "scans/blended/all_sessions/",
  rgb_file_index = 37:72,
  session        = "2022_02"
)
```

---

## Saving output

Pass `output_path` to save the result as an `.RData` file (the object is
named `root.depth.metrics` inside the file). The directory is created if it
does not exist.

```{r save}
result <- root_depth_metrics(
  path.seg    = "scans/segmented/2022_02/",
  path.skl    = "scans/skeleton/2022_02/",
  session     = "2022_02",
  output_path = "output/root_metrics_2022_02.RData"
)

# Reload later:
# load("output/root_metrics_2022_02.RData")
# head(root.depth.metrics)
```

---

## Progress and fault tolerance

By default (`verbose = TRUE`) the function prints a progress line after each
image showing per-image time, cumulative elapsed time, estimated remaining
time, and predicted clock-time of completion:

```
[RootScanR] [3/48] T042 | img: 12s | elapsed: 38s | remaining: ~546s | done ~14:22
```

If a metric block fails for a single image (e.g. a corrupted skeleton file),
it is skipped with a `[RootScanR] SKIPPED` message and the corresponding
columns are filled with `NA`. Processing always continues. Images that cannot
be loaded at all are skipped entirely and listed in a warning at the end.

To suppress all progress output:

```{r quiet}
result <- root_depth_metrics(
  path.seg = "scans/segmented/2022_02/",
  path.skl = "scans/skeleton/2022_02/",
  verbose  = FALSE
)
```

---

## Example: visualising depth profiles

```{r plot}
library(ggplot2)

# Root length density by depth, one line per tube
ggplot(result, aes(x = depth, y = rootlength.density, colour = Tube)) +
  geom_line() +
  geom_point() +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(
    title = "Root Length Density by Depth",
    x     = "Soil Depth (cm)",
    y     = "Root Length Density (cm root / cm² area)"
  )

# Average root diameter by depth
ggplot(result, aes(x = depth, y = avg.diameter)) +
  geom_smooth(method = "loess", se = TRUE) +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(
    title = "Average Root Diameter with Depth",
    x     = "Soil Depth (cm)",
    y     = "Average Diameter (cm)"
  )

# RWDI distribution across tubes
ggplot(result |> dplyr::distinct(Tube, rwdi), aes(x = rwdi)) +
  geom_histogram(bins = 15, fill = "steelblue4", colour = "white") +
  theme_minimal() +
  labs(
    title = "Root Weight Distribution Index (RWDI)",
    x     = "RWDI",
    y     = "Count"
  )
```

---

## Full example with all options

```{r full}
tube_meta <- read.csv("metadata/tube_metadata_2022_02.csv")

result <- root_depth_metrics(

  # Paths
  path.seg       = "scans/segmented/2022_02/",
  path.skl       = "scans/skeleton/2022_02/",
  path.rgb       = "scans/blended/2022_02/",

  # Per-image metadata
  insertion_angles = tube_meta$angle,
  soil_starts      = tube_meta$soil_row,
  tube_names       = tube_meta$tube_id,
  session          = "2022_02",

  # Geometry
  dpi               = 300,
  tube_diameter_cm  = 7,
  depth_interval_cm = 5,
  flat_geometry     = FALSE,

  # Core metrics
  calc_root_pixels    = TRUE,
  calc_root_length    = TRUE,
  calc_diameter_stats = TRUE,

  # Extended metrics
  calc_diameter_quantiles = TRUE,
  calc_color_metrics      = TRUE,
  calc_root_angles        = TRUE,
  calc_landscape_metrics  = FALSE,   # enable if you need patch metrics

  # Derived metrics
  calc_density_metrics      = TRUE,
  calc_distribution_indices = TRUE,
  calc_advanced_metrics     = TRUE,

  # Diameter thresholds (for quantile block)
  diameter_thresholds     = c(0.2, 0.5, 1.0),
  diameter_threshold_unit = "mm",

  # Output
  output_path = "output/root_metrics_2022_02.RData",
  verbose     = TRUE
)
```

---

## What to read next

- [Minirhizotron Scans](MinirhizotronScans_vignettes.html) — step-by-step
  explanation of each underlying function
- [Flatbed Scans](FlatBedScans_vignettes.html) — trait extraction from
  flatbed scanner images (no depth dimension)
- [Rotation Bias](Rotation_Bias_vignettes.html) — correcting for tube
  rotation artefacts before analysis
- [Function reference](../reference/index.html) — full documentation for
  every exported function
