---
title: "Analyzing Root Traits from Flatbed Scans with RootScanR in R"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Root Traits from Flatbed Scans with RootScanR in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Analyzing Root Systems from Flatbed Scans

## Introduction

Flatbed scanning captures root morphology in a 2D plane with high resolution
and controlled lighting. Unlike minirhizotron images, flatbed scans have no
depth dimension — the full root system is visible at once and traits are
computed globally rather than per depth bin.

This vignette covers the individual-function workflow for flatbed images. If
you are working with minirhizotron tubes and want a single-call batch approach,
see the [Batch Processing](BatchProcessing_vignette.html) vignette instead.

> **Prerequisite**: RootScanR works with already-segmented images (root = 1,
> background = 0). Segmentation must be done beforehand using
> [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or
> [RootPainter](https://github.com/Abe404/root_painter).

## Installation

```{r, eval=FALSE}
# install.packages("remotes")
# remotes::install_github("jcunow/RootScanR")

library(RootScanR)
library(terra)
library(tidyverse)
```

---

## Workflow

### 1. Load images

`load_flexible_image()` accepts rasters, arrays, file paths, and most common
image formats. Set `binarize = TRUE` to force the image to 0/1 values.

```{r, eval=FALSE}
# Segmented image (binary: root = 1, background = 0)
seg <- load_flexible_image(
  "path/to/segmented_scan.tif",
  output_format = "spatrast",
  normalize     = TRUE,
  binarize      = TRUE,
  select.layer  = 2        # RootDetector: layer 2 is the root channel
)

# Original RGB scan for colour analysis
rgb <- load_flexible_image(
  "path/to/rgb_scan.tif",
  output_format = "spatrast",
  normalize     = FALSE
)

plot(seg, main = "Segmented scan")
plot(rgb, main = "RGB scan")
```

---

### 2. Skeletonize

`skeletonize_image()` reduces roots to single-pixel-wide centrelines. This is
required for root length (Kimura method) and diameter estimation.

```{r, eval=FALSE}
skl <- skeletonize_image(
  seg,
  methods      = c("ZhangSuen", "GuoHall"),  # ensemble for robustness
  select.layer = 2,
  verbose      = TRUE
)

plot(skl, main = "Skeleton")
```

Available methods: `"ZhangSuen"`, `"GuoHall"`, `"MAT"` (Medial Axis
Transform), `"SteepestAscend"`. When multiple methods are supplied the results
are combined.

---

### 3. Root length

`root_length()` uses Kimura's formula, which weights orthogonal and diagonal
skeleton segments differently to approximate true path length.

```{r, eval=FALSE}
rl <- root_length(skl, unit = "cm", dpi = 300, select.layer = NULL)
cat("Total root length:", round(rl, 2), "cm\n")
```

---

### 4. Root diameter

`root_diameter()` estimates local root width from the distance transform. It
returns a diameter raster and the raw diameter values.

```{r, eval=FALSE}
diam_result <- root_diameter(
  seg,
  skeleton.img    = skl,
  skeleton_method = "MAT",
  select.layer    = NULL,
  unit            = "cm"
)

# Summary statistics
diam_vals <- terra::values(diam_result$diameter_rast, na.rm = TRUE)
cat(sprintf(
  "Diameter — mean: %.3f cm  SD: %.3f cm  max: %.3f cm\n",
  mean(diam_vals), sd(diam_vals), max(diam_vals)
))

# Distribution
hist(diam_vals,
     breaks = 40,
     xlab   = "Diameter (cm)",
     main   = "Root diameter distribution",
     col    = "steelblue4")
```

For fine/coarse root separation, `modal_peaks()` can find diameter modes in
the distribution:

```{r, eval=FALSE}
peaks <- modal_peaks(
  diam_vals,
  display_type         = "density",
  prominence_threshold = length(diam_vals) / sqrt(length(diam_vals)),
  mclust               = FALSE
)
cat("Diameter peaks at (cm):", paste(round(peaks$peak_x, 3), collapse = ", "), "\n")
```

---

### 5. Pixel counts and area coverage

```{r, eval=FALSE}
root_px <- count_pixels(seg)
void_px <- count_pixels(abs(seg - 1))

root_area_pct <- root_px / (root_px + void_px) * 100
cat(sprintf("Root area coverage: %.2f %%\n", root_area_pct))

# Root length density (cm root per cm² scan area)
scan_area_cm2 <- (root_px + void_px) / (300 / 2.54)^2
rl_density    <- rl / scan_area_cm2
cat(sprintf("Root length density: %.4f cm / cm²\n", rl_density))
```

---

### 6. Branching structure

`detect_skeleton_points()` classifies skeleton pixels into endpoints (root
tips) and branching points (forks).

```{r, eval=FALSE}
pts <- detect_skeleton_points(skl, select.layer = 2)

n_tips    <- count_pixels(pts$endpoints)
n_forks   <- count_pixels(pts$branching_points)
branch_freq <- n_forks / rl * 100   # branching points per 100 cm

cat(sprintf("Root tips: %d\n", n_tips))
cat(sprintf("Branching points: %d\n", n_forks))
cat(sprintf("Branching frequency: %.1f per 100 cm\n", branch_freq))
```

---

### 7. Root-level landscape metrics

`root_scape_metrics()` applies landscape ecology metrics to the binary root
image, treating roots as "patches". Useful for characterising spatial
complexity and connectivity.

```{r, eval=FALSE}
lsm <- root_scape_metrics(
  img     = seg,
  indexD  = NA,          # no depth index for flatbed scans
  metrics = c("lsm_c_ca", "lsm_c_np", "lsm_c_enn_mn",
              "lsm_c_area_mn", "lsm_l_ent")
)
print(lsm)
```

---

### 8. Colour analysis

`tube_coloration()` extracts mean chromatic coordinates and HSV values from
an RGB image. On flatbed scans this characterises overall root pigmentation.

```{r, eval=FALSE}
# Root pixels only
rgb_roots <- rgb
rgb_roots[seg == 0] <- NA
root_colour <- tube_coloration(rgb_roots)
print(root_colour)

# Background (soil / petri dish) pixels
rgb_bg <- rgb
rgb_bg[seg == 1] <- NA
bg_colour <- tube_coloration(rgb_bg)
print(bg_colour)
```

---

### 9. Collecting results

```{r, eval=FALSE}
results <- data.frame(
  sample_id          = "my_scan",
  root_length_cm     = rl,
  root_area_pct      = root_area_pct,
  rl_density_cm_cm2  = rl_density,
  n_tips             = n_tips,
  n_forks            = n_forks,
  branching_freq     = branch_freq,
  mean_diameter_cm   = mean(diam_vals),
  sd_diameter_cm     = sd(diam_vals)
)

write.csv(results, "flatbed_results.csv", row.names = FALSE)
```

---

## What to read next

- [Minirhizotron Scans](MinirhizotronScans_vignettes.html) — adds depth
  mapping and per-bin analysis to the individual-function workflow
- [Batch Processing](BatchProcessing_vignette.html) — wraps depth-resolved
  analysis into a single function call
- [Function reference](../reference/index.html) — full documentation for
  every exported function
