---
title: "Analyzing Root Traits from Flatbed Scans with Rootopia in R"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Root Traits from Flatbed Scans with Rootopia in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Rootopia + terra are hard dependencies, so the segmentation-based workflow
# below runs at build time against the bundled `flatbed_scan_example`. The
# displayed `load_flexible_image("path/to/...")` chunk stays eval=FALSE — that
# is how a user supplies their own files.
library(Rootopia)
library(terra)

# Faithful preview helper (vignette-local; not part of the package).
# The bundled scan is several thousand pixels wide. Shrinking that to fit an
# html_vignette figure forces the graphics device to drop one-pixel-wide roots
# and skeletons. We therefore show two views per step: a full-image overview
# for spatial context, and a centred native-resolution crop where thin features
# survive at (near) 1:1. `frac` is the fraction of each axis kept in the crop.
show_scan <- function(x, main = "", frac = 0.1, ...) {
  terra::plot(x, maxcell = Inf, main = paste0(main, " - overview"), ...)
  ev <- as.vector(terra::ext(x))            # xmin, xmax, ymin, ymax
  cx <- (ev[1] + ev[2]) / 2; cy <- (ev[3] + ev[4]) / 2
  hw <- frac * (ev[2] - ev[1]) / 2
  hh <- frac * (ev[4] - ev[3]) / 2
  terra::plot(terra::crop(x, terra::ext(cx - hw, cx + hw, cy - hh, cy + hh)),
              maxcell = Inf, main = paste0(main, " - native crop"), ...)
}
```

# Analyzing Root Systems from Flatbed Scans

## Introduction

Flatbed scanning captures root morphology in a 2D plane with high resolution
and controlled lighting. Unlike minirhizotron images, flatbed scans have no
depth dimension — the full root system is visible at once and traits are
computed globally rather than per depth bin.

This vignette covers the individual-function workflow for flatbed images. If
you are working with minirhizotron tubes and want a single-call batch approach,
see the [Batch Processing](BatchProcessing_vignette.html) vignette instead.

> **Prerequisite**: Rootopia works with already-segmented images (root = 1,
> background = 0). Segmentation must be done beforehand using
> [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or
> [RootPainter](https://github.com/Abe404/root_painter).

## Installation

```{r, eval=FALSE}
# install.packages("remotes")
# remotes::install_github("jcunow/Rootopia")

library(Rootopia)
library(terra)
library(tidyverse)
```

---

## Workflow

### 1. Load images

`load_flexible_image()` accepts rasters, arrays, file paths, and most common
image formats. The `scale` argument controls value rescaling: `"binary"` forces
0/1, `"to_01"` maps 0-255 down to 0-1, `"to_255"` maps 0-1 up to 0-255, and
`"none"` leaves values untouched. This is how you load your own files:

```{r, eval=FALSE}
# Segmented image (binary: root = 1, background = 0)
seg <- load_flexible_image(
  "path/to/segmented_scan.tif",
  output_format = "spatrast",
  scale         = "binary",
  select_layer  = 2        # RootDetector: layer 2 is the root channel
)

# Original RGB scan for color analysis
rgb <- load_flexible_image(
  "path/to/rgb_scan.tif",
  output_format = "spatrast",
  scale         = "none"
)
```

For this vignette we use the bundled example flatbed scan so every step below
runs. It is a 3-band segmentation; layer 2 is the root channel.

```{r}
data(flatbed_scan_example)
# Same call as above, but pointed at the bundled object instead of a file path:
# select_layer = 2 takes the root channel, scale = "binary" maps it to 0/1.
seg <- load_flexible_image(flatbed_scan_example, output_format = "spatrast",
                           scale = "binary", select_layer = 2)

show_scan(seg, main = "Segmented flatbed scan")
```


---

### 2. Clean-image
```{r}
seg_clean = clean_image(seg,
                        max_hole_size = 10, 
                        max_artifact_size = 400)

show_scan(seg_clean)

```




---

### 3. Skeletonize

`skeletonize_image()` reduces roots to single-pixel-wide centerlines. This is
required for root length (Kimura method) and diameter estimation.

```{r}
# `seg` is already the single root layer, so no select_layer needed here
skl <- skeletonize_image(seg, verbose = FALSE)

show_scan(skl, main = "Skeleton", frac = 2)
```

`skeletonize_image()` uses a LUT-based Zhang-Suen thinning algorithm to
reduce the segmented mask to one-pixel-wide centerlines.

---

### 4. Root length

`root_length()` uses Kimura's formula, which weights orthogonal and diagonal
skeleton segments differently to approximate true path length.

```{r}
rl <- root_length(skl, unit = "cm", dpi = 300, select_layer = NULL,
                  show_messages = FALSE)
cat("Total root length:", round(rl, 2), "cm\n")
```

---

### 5. Root diameter

`root_diameter()` estimates local root width from the distance transform. It
returns a diameter raster and the raw diameter values.

```{r}
diam_result <- root_diameter(
  seg,
  skeleton_img    = skl,
  skeleton_method = "MAT",
  select_layer    = NULL,
  unit            = "cm"
)

# Summary statistics
diam_vals <- terra::values(diam_result$diameter_rast, na.rm = TRUE)
cat(sprintf(
  "Diameter — mean: %.3f cm  SD: %.3f cm  max: %.3f cm\n",
  mean(diam_vals), sd(diam_vals), max(diam_vals)
))

# Distribution
hist(diam_vals,
     breaks = 40,
     xlab   = "Diameter (cm)",
     main   = "Root diameter distribution",
     col    = "steelblue4")
```

For fine/coarse root separation, `modal_peaks()` can find diameter modes in
the distribution:

```{r, eval=FALSE}
peaks <- modal_peaks(
  diam_vals,
  display_type         = "density",
  prominence_threshold = length(diam_vals) / sqrt(length(diam_vals)),
  mclust               = FALSE
)
cat("Diameter peaks at (cm):", paste(round(peaks$peak_x, 3), collapse = ", "), "\n")
```

---

### 6. Pixel counts and area coverage

```{r}
root_px <- count_pixels(seg)
void_px <- count_pixels(abs(seg - 1))

root_area_pct <- root_px / (root_px + void_px) * 100
cat(sprintf("Root area coverage: %.2f %%\n", root_area_pct))

# Root length density (cm root per cm² scan area)
scan_area_cm2 <- (root_px + void_px) / (300 / 2.54)^2
rl_density    <- rl / scan_area_cm2
cat(sprintf("Root length density: %.4f cm / cm²\n", rl_density))
```

---

### 7. Branching structure

`detect_skeleton_points()` classifies skeleton pixels into endpoints (root
tips) and branching points (forks).

```{r}
# `skl` is a single-layer skeleton, so no select_layer is needed
pts <- detect_skeleton_points(skl, select_layer = NULL)

n_tips    <- count_pixels(pts$endpoints)
n_forks   <- count_pixels(pts$branching_points)
branch_freq <- n_forks / rl * 100   # branching points per 100 cm

# count_pixels() returns a numeric, so coerce to integer for %d formatting
cat(sprintf("Root tips: %d\n", as.integer(n_tips)))
cat(sprintf("Branching points: %d\n", as.integer(n_forks)))
cat(sprintf("Branching frequency: %.1f per 100 cm\n", branch_freq))
```

---

### 7b. Branch order (main axis vs. laterals)

`branch_order_map()` goes one step further than `detect_skeleton_points()`:
it builds a full segment graph from the skeleton and classifies every
segment by **branch order** — the thickest, most central root is order 1
(the main axis), its laterals are order 2, their laterals order 3, and so
on. Because flatbed scans have no depth dimension, this is computed once for
the whole image.

```{r, eval=FALSE}
order_res <- branch_order_map(
  skel  = skl,    # skeleton from step 2
  mask  = seg,    # filled root mask, same grid — used for diameters
  order = "branch_order",
  unit  = "cm",
  dpi   = 300
)

# One row per order class: total length, mean diameter, tips, branch points
order_res$summary

# Rasterised branch-order classes, aligned to the skeleton
show_scan(order_res$class_map, main = "Branch order")
```

`order_metrics()` can split the architecture into the main root(s) versus
all laterals (selected by diameter, so it works regardless of how many order
classes were found):

```{r, eval=FALSE}
main_vs_lateral <- order_metrics(order_res, focal = "thickest")
print(main_vs_lateral)

# Fraction of total root length that is lateral (order >= 2)
lateral_length_fraction <- main_vs_lateral$length_fraction[main_vs_lateral$group == "rest"]
```

`order_metrics()` also accepts a numeric `focal` to split one specific order
class off from the rest:

```{r, eval=FALSE}
# Per-order-class table (same as order_res$summary)
order_metrics(order_res)

# Order 1 (main root) vs. everything else
order_metrics(order_res, focal = 1)
```

To check the classification visually, write a color-coded overlay PNG, or
re-plot a sub-window at native resolution for QC:

```{r, eval=FALSE}
order_res2 <- branch_order_map(
  skel          = skl,
  mask          = seg,
  order         = "branch_order",
  unit          = "cm",
  dpi           = 300,
  overlay_png   = "branch_order_overlay.png",
  keep_segments = TRUE
)

# Zoom into a 200x200 px window at 3x magnification
plot_order_window(
  order_res2$edges, skl,
  r_range = c(100, 300), c_range = c(100, 300),
  scale = 3, file = "branch_order_window.png"
)
```

> **For more control** over crossing resolution, pruning of weak tips, and the
> continuation rule used to group segments into roots, see
> `?root_graph_pipeline` (the engine behind `branch_order_map()`).

---

### 8. Root-level landscape metrics

`root_scape_metrics()` applies landscape ecology metrics to the binary root
image, treating roots as "patches". Useful for characterising spatial
complexity and connectivity.

```{r, eval=FALSE}
lsm <- root_scape_metrics(
  img     = seg,
  index_d  = NA,          # no depth index for flatbed scans
  metrics = c("lsm_c_ca", "lsm_c_np", "lsm_c_enn_mn",
              "lsm_c_area_mn", "lsm_l_ent")
)
print(lsm)
```

---

### 9. Color analysis

`tube_coloration()` extracts mean chromatic coordinates and HSV values from
an RGB image. On flatbed scans this characterises overall root pigmentation.

```{r, eval=FALSE}
# Root pixels only
rgb_roots <- rgb
rgb_roots[seg == 0] <- NA
root_color <- tube_coloration(rgb_roots)
print(root_color)

# Background (soil / petri dish) pixels
rgb_bg <- rgb
rgb_bg[seg == 1] <- NA
bg_color <- tube_coloration(rgb_bg)
print(bg_color)
```

---

### 10. Collecting results

```{r, eval=FALSE}
results <- data.frame(
  sample_id          = "my_scan",
  root_length_cm     = rl,
  root_area_pct      = root_area_pct,
  rl_density_cm_cm2  = rl_density,
  n_tips             = n_tips,
  n_forks            = n_forks,
  branching_freq     = branch_freq,
  mean_diameter_cm   = mean(diam_vals),
  sd_diameter_cm     = sd(diam_vals),
  lateral_length_fraction = lateral_length_fraction,
  n_root_orders      = max(order_res$edges$branch_order, na.rm = TRUE)
)

write.csv(results, "flatbed_results.csv", row.names = FALSE)
```

---

## What to read next

- [Minirhizotron Scans](MinirhizotronScans_vignettes.html) — adds depth
  mapping and per-bin analysis to the individual-function workflow
- [Batch Processing](BatchProcessing_vignette.html) — wraps depth-resolved
  analysis into a single function call
- [Function reference](../reference/index.html) — full documentation for
  every exported function
