---
title: "Minirhizotron Scans Analysis with RootScanR in R"
author: "Johannes Cunow"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Minirhizotron Scans Analysis with RootScanR in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Minirhizotron Scans Analysis with RootScanR

## Introduction

This vignette walks through the individual-function workflow for analysing
minirhizotron images — loading, calibrating, depth-mapping, and extracting
root traits step by step. It is aimed at users who want full control over each
processing stage or who are working on a specific subset of functions.

> **New to RootScanR?** If you want to process a whole set of images in one
> go, the [Batch Processing](BatchProcessing_vignette.html) vignette wraps
> this entire workflow into a single `root_depth_metrics()` call. Start there.

> **Prerequisite**: RootScanR works with already-segmented images.
> Segmentation must be done beforehand using
> [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or
> [RootPainter](https://github.com/Abe404/root_painter).

## Installation

```{r, eval=FALSE}
# install.packages("remotes")
# remotes::install_github("jcunow/RootScanR")

library(RootScanR)
library(terra)
library(tidyverse)
```

---

## Workflow overview

1. Load images
2. (Optional) Rotation censor — align field of view across sessions
3. Create depth map
4. Bin depths
5. Extract root traits per depth bin
6. Compute landscape and colour metrics
7. Derive distribution indices

---

### 1. Load images

`load_flexible_image()` accepts SpatRasters, arrays, matrices, and file paths.

```{r, eval=FALSE}
# Segmented image from RootDetector or RootPainter
seg <- load_flexible_image(
  "path/to/segmented_image.tif",
  output_format = "spatrast",
  normalize     = FALSE,
  binarize      = FALSE,   # keep original values; binarize later per layer
  select.layer  = NULL
)

# RGB scan for colour analysis
rgb <- load_flexible_image(
  "path/to/rgb_image.tif",
  output_format = "spatrast"
)

# Package example data
data(seg_Oulanka2023_Session01_T067)
seg <- terra::rast(seg_Oulanka2023_Session01_T067)
```

---

### 2. Rotation censoring (optional, recommended for multi-session studies)

When the same tube is scanned across sessions, the scanner may have rotated
slightly. Use `rotation_censor()` to crop each image to the rows that are
present in every session, so root counts are directly comparable. See the
[Rotation Bias](Rotation_Bias_vignettes.html) vignette for how to estimate
the rotation shift between sessions.

```{r, eval=FALSE}
# Crop to a 1800-pixel-wide window centred on the estimated rotation centre.
# Here we use half the image width as a simple default.
r0 <- round(dim(seg)[1] / 2, 0)

seg_censored <- rotation_censor(
  seg,
  center.offset  = r0,
  fixed.rotation = TRUE,
  fixed.width    = 1800
)
```

---

### 3. Create the depth map

`create_depthmap()` assigns a soil depth value (cm) to every pixel based on
tube geometry (diameter, tilt angle, DPI) and a sinusoidal correction for
the curvature of cylindrical tubes.

**Important parameters:**

| Parameter | Description |
|---|---|
| `sinoid` | `TRUE` for cylindrical minirhizotron tubes; `FALSE` for flat windows |
| `tube.thicc` | Inner tube diameter in cm (typically 4.4 or 7 cm) |
| `tilt` | Insertion angle from vertical in degrees (often 30–45°) |
| `dpi` | Scanner resolution |
| `start.soil` | Depth offset in cm — shifts the zero point. Provide from field calibration data. |
| `center.offset` | Fractional position of the rotational centre (0–1). 0.5 = centred. |

```{r, eval=FALSE}
# Use layer 1 as spatial template; actual root layer is separate
seg_template <- seg[[1]]

depth_map <- create_depthmap(
  img           = seg_template,
  sinoid        = TRUE,
  tube.thicc    = 7,       # cm — measure your own tube
  tilt          = 45,      # degrees
  dpi           = 300,
  start.soil    = 2.9,     # cm — from in-situ calibration
  center.offset = 0.5      # 0 = top of tube, 1 = bottom
)

# Align the depth map extent to the segmented image
depth_map <- terra::flip(terra::t(depth_map))
terra::ext(depth_map) <- terra::ext(seg_template)

plot(depth_map, main = "Depth map (cm)")
```

> **On `start.soil`**: accurate depth attribution requires knowing where
> the soil surface is in the image. In-situ calibration (marking the tube at
> installation) is the most reliable approach. Without calibration data, use
> `start.soil = 0` and interpret depths as relative to the top of the scan.

---

### 4. Bin depths

`binning()` rounds continuous depth values to discrete intervals. The bin
width (`nn`) should match your analysis scale — 5 cm is a common choice.

```{r, eval=FALSE}
depth_bins <- binning(depthmap = depth_map, nn = 5, round.option = "rounding")
```

---

### 5. Extract root traits per depth bin

#### 5a. Root pixels and void pixels

```{r, eval=FALSE}
# Select the root channel (layer 2 in RootDetector output)
root_layer <- terra::subset(seg, 2)
root_layer <- load_flexible_image(root_layer, normalize = TRUE, binarize = TRUE,
                                  output_format = "spatrast")

# Root pixels per depth bin
root_px_by_depth <- terra::zonal(root_layer, depth_bins, "sum", na.rm = TRUE)
colnames(root_px_by_depth) <- c("depth", "rootpx")

# Void pixels
void_layer <- abs(root_layer - 1)
void_px_by_depth <- terra::zonal(void_layer, depth_bins, "sum", na.rm = TRUE)
colnames(void_px_by_depth) <- c("depth", "voidpx")

depth_data <- merge(root_px_by_depth, void_px_by_depth, by = "depth")
depth_data$rootpx.density <- depth_data$rootpx /
                              (depth_data$rootpx + depth_data$voidpx) * 100
```

#### 5b. Root length (Kimura method)

```{r, eval=FALSE}
# Skeletonize first
skl <- skeletonize_image(root_layer, methods = "ZhangSuen", verbose = FALSE)

# Root length per depth bin (cm)
rl_by_depth <- terra::zonal(skl, depth_bins, "sum", na.rm = TRUE)
rl_by_depth[[2]] <- rl_by_depth[[2]] / (300 / 2.54)   # pixels → cm
colnames(rl_by_depth) <- c("depth", "rootlength")

depth_data <- merge(depth_data, rl_by_depth, by = "depth")
depth_data$rootlength.density <- depth_data$rootlength /
  ((depth_data$rootpx + depth_data$voidpx) / (300 / 2.54)^2)
```

#### 5c. Root diameter

```{r, eval=FALSE}
diam_result <- root_diameter(
  root_layer,
  skeleton.img = skl,
  unit         = "cm",
  select.layer = NULL
)
terra::ext(diam_result$diameter_rast) <- terra::ext(depth_bins)

diam_by_depth <- terra::zonal(diam_result$diameter_rast, depth_bins,
                               "mean", na.rm = TRUE)
colnames(diam_by_depth) <- c("depth", "avg.diameter")

depth_data <- merge(depth_data, diam_by_depth, by = "depth")
```

#### 5d. Zoning individual depth slices

`zoning()` masks the image to a specific depth range, useful when you want to
run per-slice analyses (landscape metrics, colour, diameter quantiles).

```{r, eval=FALSE}
# Extract a single 5 cm depth zone
zone_10cm <- zoning(
  img       = root_layer,
  mode      = "depth",
  depth_map = depth_bins,
  depth     = 10           # selects the bin closest to 10 cm
)
```

---

### 6. Landscape and colour metrics

#### Landscape metrics (spatial structure)

```{r, eval=FALSE}
# Run per depth slice inside a loop; indexD labels the output rows
lsm_list <- lapply(sort(unique(terra::values(depth_bins))), function(d) {
  slice <- zoning(root_layer, mode = "depth", depth_map = depth_bins, depth = d)
  root_scape_metrics(
    img     = slice,
    indexD  = d,
    metrics = c("lsm_c_np", "lsm_c_enn_mn", "lsm_l_ent")
  )
})
lsm_df <- do.call(rbind, lsm_list)
```

> **Note**: landscape metrics are slow — one call per depth bin per image.
> Consider enabling them only when you specifically need patch-level indices.

#### Colour metrics

```{r, eval=FALSE}
# Align RGB to segmented image
rgb_aligned <- terra::resample(rgb, root_layer, method = "bilinear")

colour_list <- lapply(sort(unique(terra::values(depth_bins))), function(d) {
  slice_seg <- zoning(root_layer,   mode = "depth", depth_map = depth_bins, depth = d)
  slice_rgb <- zoning(rgb_aligned,  mode = "depth", depth_map = depth_bins, depth = d)

  root_rgb <- slice_rgb; root_rgb[slice_seg == 0] <- NA
  bg_rgb   <- slice_rgb; bg_rgb[slice_seg == 1]   <- NA

  data.frame(
    depth = d,
    as.data.frame(tube_coloration(root_rgb)) |>
      setNames(paste0(names(tube_coloration(root_rgb)), "_root")),
    as.data.frame(tube_coloration(bg_rgb)) |>
      setNames(paste0(names(tube_coloration(bg_rgb)), "_bg"))
  )
})
colour_df <- do.call(rbind, colour_list)
```

---

### 7. Distribution indices

These tube-level summary metrics require a complete depth profile.

```{r, eval=FALSE}
# Root Weight Distribution Index — depth-weighted mean rooting depth
rwdi_val <- RWDI(w = depth_data$depth, roots = depth_data$rootlength.density)

# Root Penetration Index — how evenly roots are distributed with depth (-1 to 1)
rpi_val  <- RPI(roots = depth_data$rootlength.density, w = depth_data$depth)

# Total length density — sum over the full profile
total_ld <- sum(depth_data$rootlength.density * 5, na.rm = TRUE)

cat(sprintf("RWDI: %.2f  RPI: %.3f  Total length density: %.4f\n",
            rwdi_val, rpi_val, total_ld))
```

---

### 8. Root turnover (two-timepoint comparison)

```{r, eval=FALSE}
data(skl_Oulanka2023_Session01_T067)
data(skl_Oulanka2023_Session03_T067)
t1 <- terra::rast(skl_Oulanka2023_Session01_T067)
t2 <- terra::rast(skl_Oulanka2023_Session03_T067)

turnover <- root_turnover(
  img1   = t1,
  img2   = t2,
  method = "kimura",
  dpi    = 300,
  unit   = "cm"
)
print(turnover)
```

---

### 9. Visualise

```{r, eval=FALSE}
ggplot(depth_data, aes(x = depth, y = rootlength.density)) +
  geom_col(fill = "steelblue4") +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(x = "Soil depth (cm)", y = "Root length density (cm / cm²)",
       title = "Root length density profile")
```

---

## What to read next

- [Batch Processing](BatchProcessing_vignette.html) — the same workflow in a
  single `root_depth_metrics()` call, with fault tolerance and ETA logging
- [Flatbed Scans](FlatBedScans_vignettes.html) — trait extraction without a
  depth dimension
- [Rotation Bias](Rotation_Bias_vignettes.html) — correcting for tube rotation
  between sessions
- [Function reference](../reference/index.html) — full documentation for every
  exported function
