---
title: "Minirhizotron Scans Analysis with Rootopia in R"
author: "Johannes Cunow"
output:
  rmarkdown::html_vignette: default
vignette: >
  %\VignetteIndexEntry{Minirhizotron Scans Analysis with Rootopia in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Rootopia + terra are hard dependencies, so the workflow below can run at
# build time against the bundled example data. The displayed
# `load_flexible_image("path/to/...")` chunks stay eval=FALSE — that is how a
# user supplies their own files — while the chunks that operate on the example
# objects are eval=TRUE so the code is actually exercised.
library(Rootopia)
library(terra)
```

# Minirhizotron Scans Analysis with Rootopia

## Introduction

This vignette walks through the individual-function workflow for analyzing
minirhizotron images — loading, calibrating, depth-mapping, and extracting
root traits step by step. It is aimed at users who want full control over each
processing stage or who are working on a specific subset of functions.

> **New to Rootopia?** If you want to process a whole set of images in one
> go, the [Batch Processing](BatchProcessing_vignette.html) vignette wraps
> this entire workflow into a single `root_depth_metrics()` call. Start there.

> **Prerequisite**: Rootopia works with already-segmented images.
> Segmentation must be done beforehand using
> [RootDetector](https://github.com/ExPlEcoGreifswald/RootDetector) or
> [RootPainter](https://github.com/Abe404/root_painter).

> If a tube constitues of multiple overlapping images, consider stitching [Image Stitching](Stitching_vignette.html)


## Installation

```{r, eval=FALSE}
# install.packages("remotes")
# remotes::install_github("jcunow/Rootopia")

library(Rootopia)
library(terra)
library(tidyverse)
```

---

## Workflow overview

0. (Image Stitching & Segmentation, see: [Image Stitching](Stitching_vignette.html))
1. Load images
2. (Optional) Rotation censor — align field of view across sessions
3. Create depth map
4. Bin depths
5. Extract root traits per depth bin (pixels, length, diameter)
6. Compute landscape and color metrics per depth bin
7. Derive distribution indices
8. Root turnover (two-timepoint comparison)

> **Per depth bin** is the recurring theme below. Two patterns cover every
> trait:
>
> * **Whole profile in one call** with `terra::zonal()` — fast, for traits that
>   reduce to a per-zone sum or mean (pixel counts, mean diameter).
> * **One slice at a time** with `depth_zoning()` — for traits whose function needs a
>   whole image (root length, landscape metrics, color). You compute the trait
>   for a single depth, then wrap that in a loop to cover the full profile.
>
> Section 5 introduces both and shows how to scale a single-slice computation
> to every depth bin.

---

### 1. Load images

`load_flexible_image()` accepts SpatRasters, arrays, matrices, and file paths.
This is how you load your own files:

```{r, eval=FALSE}
# Segmented image from RootDetector or RootPainter
seg_img <- load_flexible_image(
  "path/to/segmented_image.tif",
  output_format = "spatrast",
  normalize     = FALSE,
  binarize      = TRUE,
  select_layer  = NULL
)

# RGB scan for color analysis
rgb <- load_flexible_image(
  "path/to/rgb_image.tif",
  output_format = "spatrast",
  normalize = FALSE,
  binarize = FALSE
)
```

For this vignette we use the bundled example data so every step below runs:

```{r}
# 3-band RootDetector output; layer 2 is the root channel
data(seg_Oulanka2023_Session01_T067)
seg_img    <- terra::rast(seg_Oulanka2023_Session01_T067)

# RootDetector layers are 0/255; binarize the root channel to 0/1 so that
# pixel sums and the void mask (abs(root_layer - 1)) are meaningful.
root_layer <- terra::ifel(seg_img[[2]] > 0, 1, 0)

# RGB scan (a different session of the same tube, for the color demo)
data(rgb_Oulanka2023_Session03_T067)
rgb <- terra::rast(rgb_Oulanka2023_Session03_T067)

show_root_image(root_layer, main = "Root layer (binary)")
```

---

### 2. Rotation censoring (optional, recommended for multi-session studies)

When the same tube is scanned across sessions, the scanner may have rotated
slightly. Use `rotation_censor()` to crop each image to the rows that are
present in every session, so root counts are directly comparable. See the
[Rotation Bias](Rotation_Bias_vignettes.html) vignette for how to estimate
the rotation shift between sessions.

```{r, eval=FALSE}
# Crop to a 1800-pixel-wide window centered on the estimated rotation center.
# Here we use half the image width as a simple default. If you have multiple
# sessions, consider estimating the rotation shift between sessions with
# `estimate_rotation_shift()` and adjust the center offset accordingly.

r0 <- round(dim(seg_img)[1] / 2, 0)

seg_censored <- rotation_censor(
  seg_img,
  center_offset  = r0,
  fixed_rotation = TRUE,
  fixed_width    = 1800
)
```

> Rotation censoring changes the image dimensions, so apply it **before** the
> depth map and keep every downstream raster on the censored grid. The rest of
> this vignette uses the uncensored example image.

---

### 3. Create the depth map

`create_depthmap()` assigns a soil depth value (cm) to every pixel based on
tube geometry (diameter, tilt angle, DPI) and a sinusoidal correction for
the curvature of cylindrical tubes.

**Important parameters:**

| Parameter | Description |
|---|---|
| `sinoid` | `TRUE` for cylindrical minirhizotron tubes; `FALSE` for flat windows |
| `tube_thicc` | Inner tube diameter in cm (typically 4.4 or 7 cm) |
| `tilt` | Insertion angle from vertical in degrees (often 30–45°) |
| `dpi` | Scanner resolution |
| `start_soil` | Depth offset in cm — shifts the zero point. Provide from field calibration data. |
| `center_offset` | Fractional position of the rotational center (0–1). 0.5 = centered. |

```{r}
# `mask` flags foreign objects (e.g. tube edges / tape) to exclude from depth
# attribution. Here it is derived from the RootDetector layers.
mask <- seg_img[[1]] - seg_img[[2]]
mask[mask == 255] <- NA

depth_map <- create_depthmap(
  img           = seg_img,
  mask          = mask,
  sinoid        = TRUE,
  tube_thicc    = 7,       # cm — measure your own tube
  tilt          = 45,      # degrees
  dpi           = 150,
  start_soil    = 2.9,     # cm — from in-situ calibration
  center_offset = 0.5      # 0 = top of tube, 1 = bottom
)

# create_depthmap() returns the map on its own grid; align it to the segmented
# image so depth and root rasters share extent/resolution (required for
# terra::zonal() and depth_zoning() below).
depth_map <- terra::flip(terra::t(depth_map))
terra::ext(depth_map) <- terra::ext(root_layer)

show_root_image(depth_map, fun = "mean", main = "Depth map (cm)")
```

> **On `start_soil`**: accurate depth attribution requires knowing where
> the soil surface is in the image. In-situ calibration (marking the tube at
> installation) is the most reliable approach. Without calibration data, use
> `start_soil = 0` and interpret depths as relative to the top of the scan.

---

### 4. Bin depths

`binning()` rounds continuous depth values to discrete intervals. The bin
width (`nn`) should match your analysis scale — 5 cm is a common choice.

```{r}
depth_bins <- binning(depthmap = depth_map, nn = 5, round_option = "rounding")

# The set of depth bins we will iterate over for per-slice traits
depths <- sort(unique(terra::values(depth_bins, mat = FALSE)))
depths <- depths[!is.na(depths)]
depths
show_root_image(depth_bins, fun = "mean")

```

---

### 5. Extract root traits per depth bin

#### 5a. Root pixels and void pixels — whole profile with `zonal()`

`terra::zonal()` aggregates a value raster over the zones of a class raster in
one pass. This is the fast path for traits that are just a per-bin sum or mean.

```{r}
# Root pixels per depth bin
root_px_by_depth <- terra::zonal(root_layer, depth_bins, "sum", na.rm = TRUE)
colnames(root_px_by_depth) <- c("depth", "rootpx")

# Void pixels
void_layer <- abs(root_layer - 1)
void_px_by_depth <- terra::zonal(void_layer, depth_bins, "sum", na.rm = TRUE)
colnames(void_px_by_depth) <- c("depth", "voidpx")

depth_data <- merge(root_px_by_depth, void_px_by_depth, by = "depth")
depth_data$rootpx.density <- depth_data$rootpx /
                              (depth_data$rootpx + depth_data$voidpx) * 100
head(depth_data)
```

#### 5b. Root length for a *single* depth slice — `depth_zoning()` + `root_length()`

Root length (Kimura method) is computed from a skeleton image, so it needs a
whole raster rather than a per-pixel reduction. `depth_zoning()` masks the skeleton
to one depth bin (everything outside the bin becomes `NA`, the grid is kept),
and `root_length()` then measures just that slice.

```{r}
# Skeletonize the full root layer once
skl <- skeletonize_image(root_layer, verbose = FALSE)

# Traits for one specific depth slice (the 10 cm bin)
skl_10cm <- depth_zoning(skl, depth_map = depth_bins, depth = 10)
len_10cm <- root_length(skl_10cm, unit = "cm", dpi = 150, show_messages = FALSE)
len_10cm
```

#### 5c. Scale to the whole profile — loop the single-slice computation

To get the trait for every depth, wrap the single-slice code in a loop over
`depths`. The same pattern works for any function that takes a whole image.

```{r}
length_by_depth <- do.call(rbind, lapply(depths, function(d) {
  skl_d <- depth_zoning(skl, depth_map = depth_bins, depth = d)
  # Empty bins have no skeleton to measure, so their length is 0
  has_root <- terra::global(skl_d, "sum", na.rm = TRUE)[1, 1] > 0
  data.frame(
    depth      = d,
    rootlength = if (has_root) root_length(skl_d, unit = "cm", dpi = 150,
                                           show_messages = FALSE) else 0
  )
}))

depth_data <- merge(depth_data, length_by_depth, by = "depth")
# Root length density: cm root per cm² of imaged tube wall in that bin
depth_data$rootlength.density <- depth_data$rootlength /
  ((depth_data$rootpx + depth_data$voidpx) / (150 / 2.54)^2)
head(depth_data)
```

#### 5d. Root diameter — whole profile with `zonal()`

Diameter is a per-pixel value, so once you have the diameter raster you are
back in `zonal()` territory.

```{r}
diam_result <- root_diameter(
  root_layer,
  skeleton_img = skl,
  unit         = "cm",
  select_layer = NULL
)
# Align the diameter raster to the depth grid before zonal aggregation
terra::ext(diam_result$diameter_rast) <- terra::ext(depth_bins)

diam_by_depth <- terra::zonal(diam_result$diameter_rast, depth_bins,
                               "mean", na.rm = TRUE)
colnames(diam_by_depth) <- c("depth", "avg.diameter")

depth_data <- merge(depth_data, diam_by_depth, by = "depth")
head(depth_data)
```

---

### 6. Landscape and color metrics

Both of these are computed **per depth slice** with the same `depth_zoning()` loop
introduced in 5c.

#### Landscape metrics (spatial structure)

```{r, eval=FALSE}
lsm_list <- lapply(depths, function(d) {
  slice <- depth_zoning(root_layer, depth_map = depth_bins, depth = d)
  # Skip depth bins that contain no roots
  if (terra::global(slice, "sum", na.rm = TRUE)[1, 1] == 0) return(NULL)
  root_scape_metrics(
    img     = slice,
    index_d  = d,
    metrics = c("lsm_c_np", "lsm_c_enn_mn", "lsm_l_ent")
  )
})
lsm_df <- do.call(rbind, lsm_list)
```

> **Note**: landscape metrics are slow (one call per depth bin per image) and
> require the suggested `landscapemetrics` package, so this chunk is not run
> here. Enable it when you specifically need patch-level indices.

#### Color metrics

`tube_coloration()` is the color extractor here — exactly as in the
[Flatbed Scans](FlatBedScans_vignettes.html) vignette. The only extra step for
minirhizotron data is grid alignment: the RGB scan and the segmentation often
come off the scanner on slightly different grids, and per-pixel masking
(`rgb[seg == 0] <- NA`) requires identical grids. `terra::resample()` puts the
RGB onto the segmentation grid; if your RGB and segmentation already share a
grid you can skip it.

```{r, eval=FALSE}
# Align RGB to the segmented grid (skip if they already match)
rgb_aligned <- terra::resample(rgb, root_layer, method = "bilinear")

color_list <- lapply(depths, function(d) {
  slice_seg <- depth_zoning(root_layer,  depth_map = depth_bins, depth = d)
  slice_rgb <- depth_zoning(rgb_aligned, depth_map = depth_bins, depth = d)

  # Split the slice into root vs. background pixels
  root_rgb <- slice_rgb; root_rgb[slice_seg == 0] <- NA
  bg_rgb   <- slice_rgb; bg_rgb[slice_seg == 1]   <- NA

  # Skip depth bins that contain no roots (nothing to color)
  if (all(is.na(terra::values(root_rgb[[1]], mat = FALSE)))) return(NULL)

  data.frame(
    depth = d,
    setNames(tube_coloration(root_rgb), paste0(names(tube_coloration(root_rgb)), "_root")),
    setNames(tube_coloration(bg_rgb),   paste0(names(tube_coloration(bg_rgb)),   "_bg"))
  )
})
color_df <- do.call(rbind, color_list)   # NULLs from empty bins are dropped
```

> Looking for **root branching order** (main axis vs. laterals)? That analysis
> has no depth dimension, so it lives in the
> [Flatbed Scans](FlatBedScans_vignettes.html#6b-branch-order-main-axis-vs-laterals)
> vignette. You can still run it on a minirhizotron skeleton — just compute it
> once for the whole image rather than per depth bin.

---

### 7. Distribution indices

These tube-level summary metrics require a complete depth profile.

```{r}
# Mean rooting depth — depth-weighted mean
mrd_val <- MRD(w = depth_data$depth, roots = depth_data$rootlength.density)

# Root Penetration Index — how evenly roots are distributed with depth (-1 to 1)
rpi_val  <- RPI(roots = depth_data$rootlength.density, w = depth_data$depth)

# Total length density — sum over the full profile
total_ld <- sum(depth_data$rootlength.density * 5, na.rm = TRUE)

cat(sprintf("MRD: %.2f  RPI: %.3f  Total length density: %.4f\n",
            mrd_val, rpi_val, total_ld))
```

`root_accumulation()` complements these by returning the cumulative profile
itself — useful for plotting how root mass/length accumulates with depth, or
for comparing the *shape* of accumulation curves between tubes/plots.

```{r}
# Cumulative root pixel count with depth, per tube
depth_data$Tube <- "T067"   # add a grouping column if not already present

depth_data$rootpx.cumulative <- root_accumulation(
  depth_data,
  group    = "Tube",
  depth    = "depth",
  variable = "rootpx",
  stdrz    = "relative"   # "counts" (raw), "additive" (/max), or "relative" (/sum)
)

plot(depth_data$depth, depth_data$rootpx.cumulative, type = "l",
     xlab = "Depth (cm)", ylab = "Cumulative root pixel fraction",
     main = "Root accumulation with depth")
```

`stdrz = "relative"` rescales the cumulative curve to end at 1, which makes
accumulation curves directly comparable between tubes that differ in total
root abundance — only the *shape* of the depth distribution is compared.

---

### 8. Root turnover (two-timepoint comparison)

```{r}
data(skl_Oulanka2023_Session01_T067)
data(skl_Oulanka2023_Session03_T067)
t1 <- terra::rast(skl_Oulanka2023_Session01_T067)
t2 <- terra::rast(skl_Oulanka2023_Session03_T067)

turnover <- root_turnover(
  img1      = t1,
  img2      = t2,
  method    = "tc",        # total-change family
  tc_method = "kimura",    # length estimator within that family
  dpi       = 150,
  unit      = "cm"
)
print(turnover)
```

---

### 9. Visualize

```{r, eval=FALSE}
library(ggplot2)

ggplot(depth_data, aes(x = depth, y = rootlength.density)) +
  geom_col(fill = "steelblue4") +
  coord_flip() +
  scale_x_reverse() +
  theme_minimal() +
  labs(x = "Soil depth (cm)", y = "Root length density (cm / cm²)",
       title = "Root length density profile")
```

---

## What to read next

- [Batch Processing](BatchProcessing_vignette.html) — the same workflow in a
  single `root_depth_metrics()` call, with fault tolerance and ETA logging
- [Flatbed Scans](FlatBedScans_vignettes.html) — trait extraction without a
  depth dimension, including root branching order
- [Rotation Bias](Rotation_Bias_vignettes.html) — correcting for tube rotation
  between sessions
- [Function reference](../reference/index.html) — full documentation for every
  exported function
