---
title: "Minirhizotron Scans Analysis with RootScanR in R"
author: "Johannes Cunow"
output:
  rmarkdown::html_vignette: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Minirhizotron Scans Analysis with RootScanR in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Minirhizotron Scans Analysis with RootScanR

## Introduction

This vignette demonstrates how to analyze minirhizotron scans using the **RootScanR** package. Minirhizotrons are transparent tubes inserted into soil to observe root growth via scanning or imaging technology. RootScanR is designed for comprehensive analysis of these root system images, providing tools for image preparation, depth mapping, and extraction of meaningful root traits. To extract root traits, segmenting the image outside R is needed (see step 2).

> **New to RootScanR?** If you want to process a whole set of images in one go, see the
> [Batch Processing with `root_depth_metrics()`](BatchProcessing_vignette.html) vignette, which
> wraps this entire workflow into a single function call with toggleable metric groups.

## Key Functionalities

A typical workflow to analyze minirhizotron scans could look something like the following. Steps 2 and 3 depend on the scanner setup and available calibration data.

1. **Image Stitching**: Combine sequential minirhizotron images to create complete tube scans
2. **AI-Based Image Segmentation**: Work with images segmented by dedicated root detection software
3. **In-situ Calibration**: Needed to accurately map image pixels to soil depth. An alternative is shown here but requires special attention
4. **Depthmap Creation**: Generate depth information for each pixel in the root scans
5. **Root Trait Extraction**: Calculate important root metrics (length, density, branching, etc.)
6. **Spatial Root Features**: Extracting spatial-explicit root scape features e.g., entropy, root instance count, fractal dimension
7. **Soil Characterization**: Analyze soil/peat properties from scan images

## Installation

```{r, eval=FALSE}
# Install the package from GitHub
# install.packages("remotes")
# remotes::install_github("jcunow/RootScanR")

# Load the package
library(RootScanR)
library(tidyverse)  # For data manipulation and visualization
library(terra)      # For raster image handling
```

## Example Workflow

### 1. Image Stitching

**Purpose**: Combine multiple sequential images from longer minirhizotron tubes into a single composite image to enable proper depth alignment and remove overlap between images.

> **Note**: `stitch_sequential_images()` is planned for a future release and is not yet available in the current version of RootScanR. Use external tools in the meantime — a ready-to-use Python script is available at <https://github.com/jcunow/ImageStitching>.

```{r, eval=FALSE}
# Placeholder — stitch_sequential_images() is not yet exported (to be added).
# Use external tools for now — e.g., Python code: https://github.com/jcunow/ImageStitching
```

### 2. Image Segmentation Recommendations

**Purpose**: Separate roots from background in the minirhizotron images.

RootScanR works with segmented images but does not perform segmentation itself. We recommend using specialized tools:

1. **RootDetector** ([GitHub](https://github.com/ExPlEcoGreifswald/RootDetector))
   - Returns segmented and skeletonized images with different information channels
   - Can distinguish between tape, roots, and background
   - Supports root tracking for production, decay, and turnover analysis
   - Reference: Peters et al. (2023), *Scientific Reports* **13**(1)

2. **RootPainter** ([GitHub](https://github.com/Abe404/root_painter))
   - Returns segmented images
   - Uses interactive deep learning for improved segmentation
   - Reference: Smith et al. (2022), *New Phytologist* **236**(2)

**Tip**: For images with high root density, consider splitting your stitched images before segmentation and rejoining afterwards.

**Tip**: Models can be sensitive to resolution. Try to match the DPI of your training data when applying a pre-trained model.

### 3. Loading Images

**Purpose**: Import your segmented and original RGB images for analysis.

```{r, eval=FALSE}
# Load a segmented image (replace with your file path).
# The function accepts multiple different input formats.
segmented_image <- load_flexible_image(
  "path/to/segmented_image.tif",
  output_format = "spatrast",
  normalize     = FALSE,
  select.layer  = NULL,
  binarize      = FALSE
)

# Load corresponding RGB image for visual reference
rgb_image <- load_flexible_image("path/to/rgb_image.tif", output_format = "spatrast")

# Display the images
plot(rgb_image,       main = "Original RGB Image")
plot(segmented_image, main = "Segmented Image")
```

### 4. Tube Calibration

**Purpose**: Establish the relationship between image pixels and actual soil depth.

In-situ calibration is highly recommended for accurate depth attribution. If calibration data is unavailable, RootScanR provides estimation functions using tape cover as proxy:

```{r, eval=FALSE}
# Estimate soil surface position based on tape cover.
# Here, 0.5 cm of tape beyond the true soil surface is assumed to block out light.
soil_surface <- estimate_soil_surface(rgb_image, dpi = 300, tape.overlap = 0.5)
print(paste0("Estimated soil surface begins at column: ", soil_surface$soil0))

# Estimate rotation center.
# This only works for asymmetric taping around the tube with more on the upper side.
rotation_center <- estimate_rotation_center(rgb_image)
print(paste0("Estimated rotation center at row: ", rotation_center))
```

### 5. Depthmap Creation

**Purpose**: Create a map that assigns soil depth values to each pixel in the image.

```{r, eval=FALSE}
# Create a mask to exclude tape from soil analysis.
# Assuming RootDetector format where red channel includes tape.
tape_mask <- (segmented_image[[1]] - segmented_image[[2]]) / 255
tape_mask <- terra::t(tape_mask)

# Calculate normalized rotation center position
center_offset <- rotation_center / dim(segmented_image)[1]

# Create the depthmap
depth_map <- create_depthmap(
  image         = segmented_image,
  sinusoidal    = TRUE,    # Account for tube curvature
  mask          = tape_mask,
  soil_start    = soil_surface,
  center_offset = center_offset
)

# Visualize the depthmap
plot(depth_map, main = "Depth Map")
```

### 6. Depth Binning

**Purpose**: Group depths into discrete intervals for statistical analysis and visualization.

```{r, eval=FALSE}
# Create depth bins (5 cm intervals)
binned_map <- binning(depthmap = depth_map, interval = 5)

# Correct extent for plotting
terra::ext(binned_map) <- c(0, dim(binned_map)[2], 0, dim(binned_map)[1])

# Visualize binned depthmap
transposed_map <- terra::t(binned_map)
terra::plot(transposed_map, main = "Binned Depth Map (5 cm intervals)")
```

### 7. Root Skeletonization

**Purpose**: Create a single-pixel-wide representation of root structure for morphological analysis.

```{r, eval=FALSE}
# Create root skeleton using Medial Axis Transform (MAT) method.
# Select layer 2 which typically contains root information in RootDetector output.
skeleton <- skeletonize_image(
  segmented_image,
  methods      = "MAT",
  select.layer = 2,
  verbose      = TRUE
)

# Visualize skeleton
plot(skeleton, main = "Root Skeleton")
```

### 8. Root Trait Extraction

**Purpose**: Calculate quantitative root system traits for ecological analysis.

```{r, eval=FALSE}
# Detect endpoints and branching points
skeleton_points  <- detect_skeleton_points(skeleton)
root_tips_n      <- count_pixels(skeleton_points$endpoints)
branching_pts_n  <- count_pixels(skeleton_points$branching_points)
print(paste0("Number of root tips: ",       root_tips_n))
print(paste0("Number of branching points: ", branching_pts_n))

# Calculate root length using Kimura's method
rl <- root_length(skeleton, unit = "cm", dpi = 300, select.layer = 2)
print(paste0("Total root length (cm): ", rl))

# Calculate root density
root_pixels <- count_pixels(terra::subset(segmented_image, 2))
void_area   <- abs(terra::subset(segmented_image, 2) - 1)
void_pixels <- count_pixels(void_area)
root_density <- root_pixels / (root_pixels + void_pixels)
print(paste0("Root density: ", round(root_density, 3)))

# Calculate maximum rooting depth ("deep drive")
max_depth <- deep_drive(
  RootMap  = segmented_image,
  DepthMap = depth_map
)
print(paste0("Maximum rooting depth (cm): ", max_depth))

# Calculate landscape metrics for root architecture
root_metrics <- root_scape_metrics(
  img     = segmented_image,
  indexD  = 5,
  metrics = c("lsm_c_ca", "lsm_l_ent", "lsm_c_pd", "lsm_c_enn_mn")
)
print("Root architecture metrics:")
print(root_metrics)
```

### 9. Soil/Peat Characterization

**Purpose**: Analyze soil properties visible in the minirhizotron images.

```{r, eval=FALSE}
# Analyze overall tube color
tube_color_metrics <- tube_coloration(rgb_image)
print("Overall tube color metrics:")
print(tube_color_metrics)

# Create buffer around roots to analyze soil without root & rhizosphere influence
root_buffer <- create_root_buffer(segmented_image, width = 2, halo.only = FALSE)

# Create soil-only image by removing root areas
soil_image <- rgb_image - root_buffer

# Analyze soil color (use tube_coloration, not tube_color)
soil_color_metrics <- tube_coloration(soil_image)
print("Soil color metrics:")
print(soil_color_metrics)

# Extract soil texture information using GLCM (Gray-Level Co-occurrence Matrix).
# Note: analyze_soil_texture expects a raster::brick object (required by the glcm package).
soil_raster  <- raster::brick(soil_image)
soil_texture <- analyze_soil_texture(
  img.color = soil_raster,
  grays     = 7,
  window    = c(9, 9),
  metrics   = c("second_moment", "homogeneity")
)
print("Soil texture metrics:")
print(soil_texture)
```

### 10. Depth-Based Analysis

**Purpose**: Analyze how root traits change with soil depth.

```{r, eval=FALSE}
# Load root map as binary raster
root_map <- load_flexible_image(segmented_image, binarize = TRUE, output_format = "spatrast")

# Pre-compute binned depth map (reused in every loop iteration)
binned_depth <- binning(depthmap = depth_map, interval = bin_width)

bin_width <- 2
rl.df <- data.frame(
  depth      = seq(0, 30, bin_width),
  rl         = NA_real_,
  rl.density = NA_real_,
  rootpx     = NA_real_,
  voidpx     = NA_real_
)

for (i in rl.df$depth) {
  # Subset image to this depth zone
  root_zone <- zoning(
    img       = root_map,
    mode      = "depth",
    depth     = i,
    depth_map = binned_depth
  )

  # Root length
  rl.df$rl[rl.df$depth == i] <- tryCatch(
    as.numeric(root_length(root_zone, unit = "cm", dpi = 150, select.layer = 1)),
    error = function(e) {
      message("root_length failed at depth ", i, ": ", e$message)
      NA_real_
    }
  )

  # Root pixels
  rl.df$rootpx[rl.df$depth == i] <- tryCatch(
    as.numeric(count_pixels(root_zone)),
    error = function(e) {
      message("count_pixels (root) failed at depth ", i, ": ", e$message)
      NA_real_
    }
  )

  # Void pixels (inverse of roots)
  void_zone <- abs(root_zone - 1)
  rl.df$voidpx[rl.df$depth == i] <- tryCatch(
    as.numeric(count_pixels(void_zone)),
    error = function(e) {
      message("count_pixels (void) failed at depth ", i, ": ", e$message)
      NA_real_
    }
  )

  # Root length density
  total_px <- rl.df$rootpx[rl.df$depth == i] + rl.df$voidpx[rl.df$depth == i]
  rl.df$rl.density[rl.df$depth == i] <- if (!is.na(rl.df$rl[rl.df$depth == i]) && total_px > 0) {
    rl.df$rl[rl.df$depth == i] / total_px
  } else {
    NA_real_
  }
}

# Visualize results
ggplot(rl.df, aes(x = depth, y = scale(rl.density))) +
  geom_line() +
  geom_point() +
  geom_line(aes(y = scale(rootpx)),            color = "steelblue4") +
  geom_point(aes(y = scale(rootpx)),           color = "steelblue4") +
  geom_line(aes(y = scale(voidpx + rootpx) / 10), color = "darkorange") +
  geom_point(aes(y = scale(voidpx + rootpx) / 10), color = "darkorange") +
  theme_minimal() +
  labs(
    title = "Root Length by Soil Depth",
    x     = "Soil Depth (cm)",
    y     = "Scaled value (SD units)"
  )
```

## Conclusion

This vignette demonstrates the core workflows for analyzing minirhizotron images with RootScanR. For more detailed information about specific functions and additional features, please refer to the function documentation and the package GitHub repository at <https://github.com/jcunow/RootScanR>.
