---
title: "Rotation Bias and Rhythmicity Analysis with RootScanR"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rotation Bias and Rhythmicity Analysis with RootScanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Rotation Bias Analysis

## Introduction

This vignette demonstrates how to detect and correct for rotational bias in
minirhizotron setups using the **RootScanR** package. In minirhizotron studies,
the scanner tube is inserted at an angle into the soil and can rotate slightly
between sessions. Because the CI-600 and similar scanners do not cover a full
360° arc, sequential images from the same tube may not perfectly overlap.
Left uncorrected, this rotation introduces a systematic spatial bias — roots
near the scan edges are observed in some sessions but not others.

RootScanR addresses this through three calibration functions:

- `estimate_rotation_center()` — locates the rotational zero point from tape
  coverage in the image
- `estimate_rotation_shift()` — quantifies the pixel offset between two sessions
  using image correlation
- `rotation_censor()` — crops images to the shared, overlap region

A fourth function, `estimate_soil_surface()`, locates the soil surface from
tape markers; it is used to set the depth zero reference for subsequent depth
profiling.

## Installation

```{r, eval=FALSE}
# Install from GitHub
# install.packages("remotes")
# remotes::install_github("jcunow/RootScanR")

library(RootScanR)
library(terra)
```

## Example Workflow

### 1. Estimate the Rotational Center from a Single Image

`estimate_rotation_center()` detects the white adhesive tape that is attached
to the upper side of the tube. Because more tape is visible on one side, the
function can infer where the top of the tube (rotational zero) lies. It returns
a pixel row index.

```{r, eval=FALSE}
# Load a segmented scan from the package example data
data(seg_Oulanka2023_Session01_T067)
img <- terra::rast(seg_Oulanka2023_Session01_T067)

# Detect the rotation center (returns a pixel row position)
r0 <- estimate_rotation_center(img)
print(paste("Estimated rotation center (row):", r0))
```

Key parameters:

| Parameter | Description |
|---|---|
| `tape.brightness` | Brightness threshold (0–1) for classifying tape pixels |
| `search.area` | Fraction of image width to analyse (tape is near the top) |
| `nclasses` | Number of unsupervised clustering classes |
| `tape.quantile` | Quantile used to align the brightness scale |

---

### 2. Estimate the Rotation Shift Between Two Sessions

When the same tube is scanned in two different sessions, the scanner may have
rotated. `estimate_rotation_shift()` uses either cross-correlation (`"ccf"`) or
phase correlation (`"phase"`) on a shared depth window to find the pixel offset.

```{r, eval=FALSE}
data(seg_Oulanka2023_Session01_T067)
data(seg_Oulanka2023_Session03_T067)

img1 <- seg_Oulanka2023_Session01_T067  # session 1
img2 <- seg_Oulanka2023_Session03_T067  # session 3

# Phase correlation is generally more robust to brightness differences
shift <- estimate_rotation_shift(img1, img2, cor.type = "phase",
                                 fixed.depth.pixel = c(1000, 4000))

cat("Rotation shift (x, y pixels):", shift, "\n")
```

The returned vector contains the horizontal (depth) and vertical (rotation)
pixel shifts. A large vertical shift means the tube rotated substantially
between sessions.

---

### 3. Censor Image Edges to the Shared Overlap Region

Once the rotational center and shift are known, `rotation_censor()` crops each
image so that only the rows present in every session are retained. This
eliminates the non-overlapping margins and makes root counts directly comparable
across sessions.

Two modes are available:

- **`fixed.rotation = FALSE`**: cuts proportionally based on the measured offset
- **`fixed.rotation = TRUE`**: centers the crop on a specified row and forces a
  fixed output width

```{r, eval=FALSE}
data(seg_Oulanka2023_Session01_T067)
img <- terra::rast(seg_Oulanka2023_Session01_T067)

# Variable mode — trim proportionally to the measured shift
censored_var <- rotation_censor(img,
                                center.offset = 120,
                                cut.buffer    = 0.02,
                                fixed.rotation = FALSE)

# Fixed mode — crop to a 1000-pixel-wide window centred on row 220
censored_fix <- rotation_censor(img,
                                center.offset  = 220,
                                cut.buffer     = 0.02,
                                fixed.width    = 1000,
                                fixed.rotation = TRUE)

terra::plot(censored_fix, main = "Censored image (fixed rotation)")
```

> **Note on tube geometry.** The inner and outer tube diameters differ, so the
> observed root length is a slight underestimate of the true root length in the
> soil. A resize coefficient may be applied separately; `rotation_censor()`
> does not handle this correction.

---

### 4. Locate the Soil Surface

`estimate_soil_surface()` finds where soil begins by detecting the lower edge
of the adhesive tape. It returns a data frame with two columns:

- `soil0` — pixel column of the soil surface
- `tape.end` — pixel column of the tape's lower margin (with a safety buffer)

These values feed directly into `create_depthmap()` as `start.soil`.

```{r, eval=FALSE}
data(rgb_Oulanka2023_Session03_T067)
img_rgb <- rgb_Oulanka2023_Session03_T067

surface <- estimate_soil_surface(img_rgb, dpi = 150)
print(surface)

# Use soil0 to anchor the depth map at the true soil surface
cat("Soil surface at pixel column:", surface$soil0, "\n")
```

Key parameters:

| Parameter | Description |
|---|---|
| `dpi` | Scanner resolution; needed to convert the pixel position to cm |
| `tape.overlap` | Safety margin (cm) subtracted from the tape edge |
| `inverse` | Set `TRUE` if the tape is darker than the soil |
| `nclasses` | Number of clustering classes for tape detection |

---

## Rhythmicity Analysis

In some minirhizotron deployments, root observations are made at high temporal
frequency (e.g. multiple times per day). In these datasets it can be useful to
test whether root dynamics show rhythmic — for instance diel — patterns.

`rhythmicity()` fits a sine curve of the form

$$y = A \sin\!\left(\frac{2\pi}{P}(x + \phi)\right) + c$$

to a time series and tests whether the amplitude $A$ is significantly different
from zero, using your choice of three statistical tests.

### Fitting and Testing

```{r, eval=FALSE}
set.seed(42)
x <- seq(0, 48, length.out = 120)           # 48 hours, 120 observations
y <- 1.8 * sin(2 * pi / 24 * (x + 5)) +    # amplitude 1.8, 24-h period
     4 + rnorm(120, 0, 1.2)                 # baseline + noise

# --- F-test (default) ---
res_F <- rhythmicity(x, y, fix_period = 24, method = "F")
cat(sprintf("Amplitude: %.2f | Phase: %.1f h | R²: %.3f | p (F): %.4f\n",
            res_F$amplitude, res_F$phase, res_F$R2, res_F$p_value))

# --- Finite-sample likelihood ratio test ---
res_FLR <- rhythmicity(x, y, fix_period = 24, method = "FLR")
cat(sprintf("p (FLR): %.4f\n", res_FLR$p_value))

# --- Large-sample likelihood ratio test ---
res_LR <- rhythmicity(x, y, fix_period = 24, method = "LR")
cat(sprintf("p (LR): %.4f\n", res_LR$p_value))
```

### Estimating an Unknown Period

If you do not know the period in advance, set `fix_period = NULL` and provide
starting values for all four parameters. The optimizer will estimate the period
along with amplitude, phase, and offset.

```{r, eval=FALSE}
res_free <- rhythmicity(x, y,
                        fix_period = NULL,
                        parStart   = list(amp = 2, phase = 0,
                                          offset = 4, period = 20))

cat(sprintf("Estimated period: %.1f h | R²: %.3f | p: %.4f\n",
            res_free$period, res_free$R2, res_free$p_value))
```

> **Caution.** With an unknown period, the model has four free parameters and
> the likelihood surface can be multimodal. Provide ecologically motivated
> starting values (e.g. `period = 24` for a diel hypothesis) and inspect the
> fitted curve before interpreting results.

### Plotting the Fitted Curve

`rhythmicity()` does not produce plots directly, but the fitted values are
easily recovered via the underlying `fit_sine_curve()` helper (not exported;
call via `RootScanR:::fit_sine_curve()`), or reconstruct the curve from the
returned parameters:

```{r, eval=FALSE}
# Reconstruct the fitted sine from the returned parameters
x_seq  <- seq(min(x), max(x), length.out = 300)
y_fit  <- res_F$amplitude *
           sin(2 * pi / res_F$period * (x_seq + res_F$phase)) +
           res_F$offset

plot(x, y, pch = 16, col = "grey50",
     xlab = "Time (hours)", ylab = "Root growth index",
     main = "Rhythmicity fit (fixed 24-h period)")
lines(x_seq, y_fit, col = "coral", lwd = 2)
legend("topright",
       legend = sprintf("A = %.2f, R² = %.3f, p = %.4f",
                        res_F$amplitude, res_F$R2, res_F$p_value),
       bty = "n")
```

### Interpreting the Output

The returned list includes:

| Element | Description |
|---|---|
| `amplitude` | Estimated amplitude $A$ |
| `phase` | Phase shift $\phi$ (modulo period) |
| `offset` | Baseline $c$ |
| `period` | Fixed or estimated period $P$ |
| `peakTime` | Time of peak within one cycle ($P/4 - \phi \mod P$) |
| `R2` | Coefficient of determination |
| `stat` | Test statistic (F, FLR, or LR depending on `method`) |
| `p_value` | P-value for the test of $A \neq 0$ |

A significant p-value indicates that a rhythmic component explains a
non-trivial portion of the variance. Inspect $R^2$ to judge effect size — a
significant but low-$R^2$ result may reflect noise or a poorly chosen period.

---

## Connecting Rotation Bias and Rhythmicity

When rotation-censored images are used in a high-frequency monitoring setup,
the two analyses can be connected in a straightforward pipeline:

1. For each session, apply `estimate_rotation_shift()` relative to a reference
   session to obtain per-session offsets.
2. Pass each image through `rotation_censor()` to harmonise the field of view.
3. Extract root coverage or length per depth bin using `root_depth_metrics()`.
4. Test whether the resulting time series exhibits rhythmic dynamics using
   `rhythmicity()`.

This ensures that apparent rhythmic signals are not artefacts of the tube
rotating in and out of view between sessions.

---

## Further Reading

For depth profiling, trait extraction, and turnover analysis see the other
RootScanR vignettes. Full function documentation is available at
[https://jcunow.github.io/RootScanR/](https://jcunow.github.io/RootScanR/).
Source code and issue tracker: [https://github.com/jcunow/RootScanR](https://github.com/jcunow/RootScanR).
