---
title: "Special Topics: Soil Color, Soil Texture, Rhizosphere Halos, and Turnover"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Special Topics: Soil Color, Soil Texture, Rhizosphere Halos, and Turnover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
library(Rootopia)
```

This article collects a few less-obvious capabilities of **Rootopia** that
do not fit the main step-by-step tutorials but are useful for specific
questions: classifying soil/soil material by color, quantifying soil
surface texture, building a rhizosphere "halo" zone around roots, and
measuring root turnover between time points.

All examples use the bundled Oulanka 2023 example rasters, so you can run
them as-is.

# 1. Soil and material color classification

`classify_soil_rgb()` assigns every pixel of an RGB scan to a
class (e.g., dark soil, red soil, root, silver tape, coarse debris, ...) by
nearest-centroid matching in CIE LAB color space. Pixels too far from any
centroid are left "unclassified".

```{r soil-classify}
img <- terra::rast(rgb_Oulanka2023_Session03_T067)

# downsample_fact speeds up the demo; drop it for full resolution
result <- classify_soil_rgb(img, downsample_fact = 4, verbose = FALSE)

# A SpatRaster of class IDs (factor levels are the class names)
terra::plot(result$map)
```

The returned list also carries per-class statistics -- pixel counts, area
fractions, mean LAB/RGB colors, and the mean distance to the centroid:

```{r soil-metrics}
result$metrics
```

## Visualizing the classification

`plot_soil_classification()` renders the class map with a legend and the
actual mean colors of each class:

```{r soil-plot}
plot_soil_classification(result)
```

## Calibrating your own centroids

The default centroids were calibrated on one scanner and site. For other
data, build your own with `build_soil_centroids()`. You supply `picks`: a
named list with one element per material class, where each element is a
matrix with 3 columns (R, G, B in 0-255). Each row is one color sample for
that class — different classes can have different numbers of rows. The
function averages each class's samples into a single LAB centroid and returns
a table in the same format as the built-in defaults, ready to pass back into
`classify_soil_rgb()`.

The simplest approach is to read representative RGB values off your scan
with a color picker (e.g. in an image viewer or QGIS) and enter them
directly:

```{r soil-centroids, eval = FALSE}
# each row is one pixel with r,g,b values
picks <- list(
  region1 = matrix(c( 28,  22,  18,
                         32,  26,  21,
                         25,  19,  15), ncol = 3, byrow = TRUE),
  region2  = matrix(c( 80,  45,  35,
                         65,  32,  31), ncol = 3, byrow = TRUE),
  region3      = matrix(c(180, 160, 130,
                       175, 155, 125,
                       185, 165, 135,
                       178, 158, 128), ncol = 3, byrow = TRUE),
  region4      = matrix(c(201, 205, 210,
                       198, 200, 205), ncol = 3, byrow = TRUE)
)

# One MAX_DIST (LAB units, ~10-30) per class -- larger admits more pixels
max_dist <- c(region1 = 14, region2 = 14, region3 = 26,
              region4 = 12)

cents  <- build_soil_centroids(picks, max_dist)   # prints diagnostics
result <- classify_soil_rgb(img, centroids = cents)
terra::plot(result$map)
```

Note: provide a class for **every** material in your scans. Because
classification is nearest-centroid, any material without its own class is
snapped into whichever defined class sits closest. `build_soil_centroids()`
prints intra-class spread and inter-class distances to help you tune
`MAX_DIST` before running the full classification.

See `?classify_soil_rgb` for the full description of the centroid table
format and the `prior`/`alpha` blending workflow for iterative refinement.

# 2. Soil surface texture and color

`analyze_soil_texture()` computes gray-level co-occurrence matrix (GLCM)
texture metrics from a color image, which can be informative on the spatial heterogeneity within a scan.

```{r soil-texture, eval = requireNamespace("glcm", quietly = TRUE)}
tex <- analyze_soil_texture(
  img,
  grays   = 12,
  window  = c(5, 5),
  metrics = c("variance", "second_moment","correlation","entropy")
)
terra::plot(tex)
```

A simple summary of overall tube color (mean RGB and a single
luminance value) is available via `tube_coloration()`:

```{r tube-coloration}
tube_coloration(img)
```




# 3. Rhizosphere halo (root buffer zone)

`create_root_buffer()` grows a buffer of a chosen width around every root
pixel -- a simple proxy for the rhizosphere / diffusion zone. With
`halo_only = TRUE` it returns only the ring around the roots (excluding the
roots themselves).

```{r root-buffer}
seg  <- load_flexible_image(seg_Oulanka2023_Session03_T067, select_layer = 2)
halo <- create_root_buffer(seg, width = 3, halo_only = TRUE, kernel = "circle")
terra::plot(halo)

halo10 <- create_root_buffer(seg, width = 10, halo_only = TRUE, kernel = "circle")
terra::plot(halo10)
halo10 <- create_root_buffer(seg, width = 10, halo_only = FALSE, kernel = "circle")
terra::plot(halo10)
```

The `kernel` argument switches between a `"circle"` (8-neighbor) and a
`"diamond"` (4-neighbor) growth shape, and `width` controls how many
dilation iterations are applied. Set `halo_only = FALSE` to return the
roots plus their buffer as a single filled mask.

# 4. Root turnover between time points

`root_turnover()` quantifies how much root material was produced, lost, or
retained. `method` selects one of two approaches:

- `"tc"` (Temporal Comparison) -- compare two segmented images from
  different sessions. The `tc_method` argument picks how root amount is
  measured: `"kimura"` (root length) or `"rootpx"` (root pixel count).
- `"dpc"` (Decay, Production, Constant) -- decompose a single multi-layer
  image produced by RootDetector into decayed, newly produced, and unchanged
  root fractions.

The bundled `TurnoverDPC_data` is a DPC-style image:

```{r turnover-dpc}
dpc <- terra::rast(TurnoverDPC_data)
root_turnover(dpc, method = "dpc")
```

For the two-timepoint comparison you would instead pass both sessions:

```{r turnover-two, eval = FALSE}
s1 <- terra::rast(skl_Oulanka2023_Session01_T067)
s3 <- terra::rast(skl_Oulanka2023_Session03_T067)
root_turnover(s1, s3, method = "tc", tc_method = "rootpx", unit = "cm", dpi = 150, select_layer = 2)
```

# See also

- `vignette("BatchProcessing_vignette")` -- end-to-end depth profiling.
- `?classify_soil_rgb`, `?build_soil_centroids`, `?plot_soil_classification`
- `?analyze_soil_texture`, `?tube_coloration`
- `?create_root_buffer`
- `?root_turnover`
