---
title: "Stitching Scan Sequences into Mosaics with Rootopia"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stitching Scan Sequences into Mosaics with Rootopia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.align = "center"
)
has_png <- requireNamespace("png", quietly = TRUE)   # demo writes temp PNGs
```

# Stitching Scan Sequences into Mosaics

Minirhizotron cameras and flatbed scanners often capture a tube (or a long
sample) as a **sequence of overlapping frames**. One function turns a folder of
those frames into one mosaic per tube:

```{r wrapper, eval = FALSE}
stitch_root_scans("path/to/scans", pattern = ".tiff", out_dir = "path/to/output")
```

It discovers the files, groups them into one sequence per tube, sorts each
group, aligns and blends the frames, and (optionally) writes one PNG per tube.
That is the only function you need to call. (The per-pair and per-sequence
engines it uses are documented at `?stitch_image_pair` and
`?stitch_image_sequence` for advanced use; `?list_tubes` and `?list_scan_files`
preview a folder.)

```{r lib, eval = FALSE}
# install.packages("remotes")
# remotes::install_github("jcunow/Rootopia")
library(Rootopia)
```

```{r lib-real, include = FALSE}
library(Rootopia)
```

---

## Any input format

Frames are loaded through `load_flexible_image()`, so the **same call works
whatever your format**: file paths (`.png`, `.jpg/.jpeg`, `.tif/.tiff`, `.bmp`),
a `terra` `SpatRaster`, a `raster` `RasterLayer`/`RasterBrick`, an `imager`
`cimg`, a `magick-image`, or a plain `matrix` (grayscale) / `array`
(`height x width x channel`). Color and grayscale both work — color is reduced
to luma for the alignment step only.

---

## What happens under the hood

For each consecutive pair of frames A (left) and B (right) within a tube, the
wrapper:

1. **Edge band** — takes a vertical band (rows `vertical_offset` to
   `vertical_offset + vertical_region`) of A's **right** `edge_width` columns and
   B's **left** `edge_width` columns.
2. **Reduce & preprocess** — converts the band to luma and applies `preprocess`
   if set.
3. **Align** — computes the **FFT phase correlation** of the two bands; the peak
   gives the shift `(dx, dy)` and a confidence `peak`.
4. **Place** — puts B at column `wA - edge_width + dx`, so
   **`overlap = edge_width - dx`**, and offsets it vertically by `dy`; the canvas
   grows to fit.
5. **Composite** — combines the overlap with `blend`.

These five steps repeat along the sequence, each frame onto the growing mosaic.
Understanding them is enough to understand every setting below.

---

## Demo: slice a real scan and stitch it back

`rgb_Oulanka2023_Session03_T067` is a real RGB tube scan. We cut it into
overlapping pieces, save them as a little "tube" folder, and let
`stitch_root_scans()` reassemble it — exactly the path your own scans take.

```{r load-real}
data("rgb_Oulanka2023_Session03_T067")

# load to an (H, W, C) array; downsampled here only to keep the vignette light
img <- load_flexible_image(rgb_Oulanka2023_Session03_T067,
                           output_format = "array", normalize = FALSE)
img <- img[seq(1, nrow(img), 3), seq(1, ncol(img), 3), , drop = FALSE]
dim(img)                                   # rows (rotation) x cols (length) x RGB
```

```{r slice}
W       <- ncol(img)
piece_w <- W %/% 3L                        # each piece ~1/3 of the length
ov      <- piece_w %/% 4L                  # overlap = 25% of a piece
step    <- piece_w - ov
starts  <- (0:((W - piece_w) %/% step)) * step
pieces  <- lapply(starts, function(s) img[, (s + 1):(s + piece_w), , drop = FALSE])
length(pieces)                             # number of overlapping frames
```

```{r write-tube, eval = has_png}
# save the pieces as one tube ("T067") in a temp folder, as if freshly scanned
d  <- file.path(tempdir(), "stitch_demo")
dir.create(d, showWarnings = FALSE)
sc <- if (max(img, na.rm = TRUE) > 1) 255 else 1
for (i in seq_along(pieces)) {
  png::writePNG(pmin(pmax(pieces[[i]] / sc, 0), 1),
                file.path(d, sprintf("Oulanka_T067_%02d.png", i)))
}
list.files(d)
```

```{r stitch-wrapper, eval = has_png}
res <- stitch_root_scans(d, pattern = ".png", edge_width = ov,
                         vertical_region = nrow(img), vertical_offset = 0,
                         report = TRUE, verbose = FALSE)
mosaic <- res$mosaics[["T067"]]
dim(mosaic)            # reconstructed length
res$report             # per-join dx, dy, peak (confidence), overlap
```

```{r show-real, eval = has_png, fig.height = 4.5}
norm01 <- function(a) a / max(a, na.rm = TRUE)         # display only
span   <- ncol(mosaic)

op <- graphics::par(mfrow = c(2, 1), mar = c(0, 0, 1.4, 0))
plot(grDevices::as.raster(norm01(img[, seq_len(span), , drop = FALSE])))
graphics::title("original scan (cropped to the stitched span)")
plot(grDevices::as.raster(norm01(mosaic)))
graphics::title("re-stitched by stitch_root_scans()")
graphics::par(op)
```

The recovered `dx` is ~0 (we sliced with identical edges) and `peak` is high — a
clean alignment — and the two panels are indistinguishable. (This chunk needs
the suggested `png` package to write the demo files.)

---

## Choosing settings

Each setting maps onto a step above.

**`edge_width`** — the band width used to align. Because
`overlap = edge_width - dx`, set it to roughly **1-2x your true overlap** (the
overlap should be between about half and all of `edge_width`). Too small misses
matching content; too large dilutes the peak.

**`vertical_region` / `vertical_offset`** — which rows are matched. Raise
`vertical_offset` to skip a header, label, or tape strip at the top.

**`direction`** — `"horizontal"` (default, frames side-by-side) or `"vertical"`
(frames stacked top-to-bottom, e.g. strips acquired *down* a tube).

**`method`** — `"phase"` (FFT phase correlation) is supported. The Python tool's
`"feature"` (SIFT/ORB) needs OpenCV and has no CRAN backend, so it errors with a
message rather than guessing.

**`preprocess`** — applied to the edge bands before correlation:

| value | effect | use when |
|---|---|---|
| `"none"` | raw luma | well-textured color scans |
| `"center"` | subtract mean | mild offset differences |
| `"hann"` | demean + Hann window | suppress FFT edge/wrap artifacts |
| `"grad"` / `"grad_norm"` | gradient magnitude | uneven lighting / vignetting, sparse texture |
| `"norm"` / `"center_norm"` | scale by SD | little effect on phase correlation alone |

**`blend` / `blend_width`** — how the overlap is combined:

| value | effect | use when |
|---|---|---|
| `"linear"` | feather (alpha 1->0) | color scans; hides exposure seams |
| `"overlay"` | second frame on top | hard seam, no averaging |
| `"overlay_first"` | first frame on top | hard seam, keep frame 1 |
| `"max"` | per-pixel brighter / union | **segmented / binary masks** |
| `"min"` | per-pixel darker | dark-feature masks |

For **segmented masks** (root = 1, background = 0) prefer `"max"`: feathering
*averages*, turning a root pixel into a fractional value (e.g. 0.5) and
**ghosting thin roots** where alignment is off by a pixel. `"max"` keeps any
root from either frame and never invents in-between values. `blend_width`
narrows the linear ramp to a centered band to reduce ghosting on color scans.

### Cheat-sheet by image type

| Your data | Suggested call |
|---|---|
| Color flatbed / RGB tube frames | `preprocess = "none"`, `blend = "linear"` |
| Uneven lighting / vignetting | `preprocess = "grad"` (or `"hann"`) |
| Segmented / binary masks | `blend = "max"`, `preprocess = "none"` |
| Strips acquired down the tube | `direction = "vertical"` |
| Low, fixed overlap | `edge_width` ~ 1-2x that overlap |

Only how you *supply* frames changes with format; the settings behave the same.

---

## Working with a real folder

```{r folder, eval = FALSE}
# preview first
list_tubes("path/to/scans", pattern = ".tiff")        # index | tube | n_frames

# stitch every tube, one PNG each
stitch_root_scans("path/to/scans", pattern = ".tiff", out_dir = "path/to/output")

# stitch a subset, pick a blend for segmented masks, and get a quality report
res <- stitch_root_scans(
  "path/to/scans", pattern = ".tiff",
  tubes  = 1:36,           # or tubes = c("T037","T040"), or tubes = "ask"
  blend  = "max",
  report = TRUE
)
res$mosaics                                  # named list, one array per tube
res$report                                   # tube | step | dx | dy | peak | overlap
aggregate(peak ~ tube, res$report, mean)     # mean confidence per tube
```

By default frames are stitched in **sorted file-name order** (see
*Naming, selection & ordering* below to change it). `stitch_root_scans()` also
accepts a plain character vector of paths.

---

## Naming, selection & ordering

**Grouping — `group_regex`.** A regular expression that pulls a tube id out of
each path; files sharing that id form one sequence. The default `"T0\\d{2}"`
(the `\\` is R string-escaping for the regex `T0\d{2}`) matches `T0` plus two
digits, e.g. `T067`. Adapt it to your names, or use `NULL` to put every file in
a single mosaic:

```{r grp, eval = FALSE}
group_regex = "T\\d+"          # T1, T67, T123
group_regex = "tube[-_]?\\d+"  # tube12 / tube_12 / tube-12
group_regex = NULL             # no grouping -> one mosaic
```

**Selecting — `select` vs `tubes`.** `select` indexes the sorted *file* list
*before* grouping; `tubes` selects whole *tubes* by index, name, or
interactively. Use `list_scan_files()` / `list_tubes()` to see the indices.

```{r sel, eval = FALSE}
stitch_root_scans("scans", pattern = ".tiff", select = 1:36)              # first 36 files
stitch_root_scans("scans", pattern = ".tiff", tubes  = 1:36)              # first 36 tubes
stitch_root_scans("scans", pattern = ".tiff", tubes  = c("T037", "T040")) # named tubes
stitch_root_scans("scans", pattern = ".tiff", tubes  = "ask")             # interactive
```

Normally you want `tubes`; `select` is for raw file ranges.

**Ordering — `order_by` / `decreasing`.** Frames within a tube are stitched in
order. The default key is the **file name**, which sorts *lexicographically* —
so unpadded numbers come out wrong (`_10` before `_2`). Use `order_by` to sort
by the frame number instead, and `decreasing = TRUE` to reverse it (e.g. frames
numbered from the bottom of the tube upward):

```{r ord, eval = FALSE}
# order by the number just before the extension, e.g. ..._02.tiff, ..._10.tiff
stitch_root_scans("scans", pattern = ".tiff", order_by = "\\d+(?=\\.)")
stitch_root_scans("scans", pattern = ".tiff", order_by = "\\d+(?=\\.)", decreasing = TRUE)
```

(Equivalently, zero-pad your frame numbers — `_02`, `_10` — so the default
name sort is already correct.)

**Output names — `out_prefix` / `out_format`.** When `out_dir` is set, each
mosaic is written to `<out_dir>/<out_prefix><tube>.<ext>`. `out_prefix` tags a
session/site so separate runs don't collide; `out_format` picks `"png"`
(default) or `"tiff"`:

```{r pre, eval = FALSE}
stitch_root_scans("scans/Oulanka2020_November", pattern = ".tiff",
                  out_dir = "mosaics", out_prefix = "Oulanka2020_Session03_",
                  out_format = "tiff")
# -> mosaics/Oulanka2020_Session03_T067.tif, ...T068.tif, ...
```

---

## Quality control

Use the report's `peak` to spot weak joins before trusting a mosaic:

```{r qc, eval = FALSE}
res <- stitch_root_scans("path/to/scans", pattern = ".tiff", report = TRUE)
res$report[order(res$report$peak), ][1:10, ]   # least-confident joins first
```

If a tube aligns poorly: bring `edge_width` closer to the real overlap, raise
`vertical_offset` past a header/tape strip, try `preprocess = "grad"` for uneven
lighting, or confirm the frames are in the intended (filename) order.
